---
title: "Explore p-value distribution problem using ASH: RNA vs. Ribo divergence"
author: "Joyce Hsiao"
date: "2015-10-06"
output: 
  html_document:
    toc: true
---


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```

In all of our interaction testings of differential divergence between molecular layers, we observe an odd distribution of p-values where the histogram exhibits a sharp decrease at the range of low p-values. 

Here we apply Matthew's ASH package to estimate the distribution under which these p-valuees are likey to be generated. 


Set up
======

```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
              bootstrap.panel = TRUE, 
              bootstrap.show.code = TRUE)
```


```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)

library(qvalue)

## Install Matthew's package
install_github("stephens999/ashr")
library(ashr)
```



Helper functions
================

Generate regression coefficients and their standard errors.

```{r}
interact2way_coef <- function (eSet) 
{
    fNames <- featureNames(eSet)
    require(nlme)
    gls.res <- lapply(1:length(fNames), function(i) {
        mat1 <- eSet[featureNames(eSet) == fNames[i], ]
        mat1aov.temp <- data.frame(cov = c(t(exprs(mat1))), seqData = mat1$seqData, 
            species = mat1$species, celline = mat1$celline)
        mat1aov.temp$species <- as.factor(mat1aov.temp$species)
        mat1aov.temp$seqData <- as.factor(mat1aov.temp$seqData)
            fit_interact_try <- tryCatch(fit_interact <- gls(cov ~ 
                species * seqData, weights = varIdent(form = ~1 | 
                seqData), data = mat1aov.temp, na.action = na.omit), 
                condition = function(c) c)
                  coef_table <- summary(fit_interact_try)$tTable
                  return(coef_table)
    })
    names(gls.res) <- fNames
    return(gls.res)
}
```



Compute regression coefficients
========================

Ash works with regression coefficients. We fit the interaction model of two main effects of species and technology and then extract the regression coefficients.

`eSetRRP.RP.Q.log2` includes log2 quantile-normalized RPKM values.


```{r, eval = FALSE}
load(file.path(rdadir,"eSetRRP.rda"))

# Human vs. Chimp
eSet_human_chimp <- eSetRRP.RP.Q.log2[ , 
          eSetRRP.RP.Q.log2$seqData != "protein" & eSetRRP.RP.Q.log2$species != "rhesus"]
riborna_coef_human_chimp <- interact2way_coef(eSet_human_chimp)

# Human vs. Rhesus
eSet_human_rhesus <- eSetRRP.RP.Q.log2[ , 
          eSetRRP.RP.Q.log2$seqData != "protein" & eSetRRP.RP.Q.log2$species != "chimp"]
riborna_coef_human_rhesus <- interact2way_coef(eSet_human_rhesus)

# Chimp vs. Rhesus
eSet_chimp_rhesus <- eSetRRP.RP.Q.log2[ , 
          eSetRRP.RP.Q.log2$seqData != "protein" & eSetRRP.RP.Q.log2$species != "human"]
riborna_coef_chimp_rhesus <- interact2way_coef(eSet_chimp_rhesus)

save(riborna_coef_human_chimp,
     riborna_coef_human_rhesus,
     riborna_coef_chimp_rhesus, file = file.path(rdadir,"riborna_coef.rda"))
```



Distribution of the p-values.

```{r}
load(file = file.path(rdadir, "riborna_coef.rda") )

pvals <- sapply(riborna_coef_human_chimp, function(per_gene) {
                per_gene[4,4]
          })

hist(pvals, main = "Human vs. Chimp")
```




ASH
===

Extract per gene regression coefficients and their standard errors.

```{r}
load(file = file.path(rdadir,"riborna_coef.rda"))

betas_df <- lapply(riborna_coef_human_chimp, function(per_gene) {
              data.frame(beta = per_gene[4, 1],
                         se = per_gene[4, 2])
})
betas_df <- do.call(rbind, betas_df)
head(betas_df)
```


Run Ash with 16 degrees of freedom.

```{r}
beta.ash <- ash(betas_df[,1], betas_df[,2], df = 16)

str(beta.ash)
```

Pi'0s

```{r}
with(beta.ash, round(fitted.g$pi[fitted.g$pi != 0], 4) ) 
with(beta.ash, fitted.g$b[fitted.g$pi != 0])

summary(betas_df$beta)
```


Regression coefficients and posterior means.

```{r}
plot(x = betas_df[,1],
     y = beta.ash$PosteriorMean,
     xlab = "Regression coefficients",
     ylab = "Posterior means")
abline(0, 1, col = "red")
abline(h = 0, v = 0, col = "blue")
```

ZeroProb.

```{r}
hist(beta.ash$ZeroProb)

sum(beta.ash$ZeroProb < .1)
```




Session information
========================


```{r}
sessionInfo()
```

