---
title: "RNA vs. Ribo divergence, regression coefficients, voom adjusted"
author: "Joyce Hsiao"
date: "2015-11-18"
output: 
  html_document:
    toc: true
---


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


## Objectives

We computed voom weights for every gene and then applied these weights to the data while
testing the null hyothesis that divergence is the same between RNA and Ribo levels, using the two approaches:

1. t-test for regression coefficients

2. Likelihood ratio model 

Results show that t-test is more conservative thatn the Likelihood ratio test in detecting number of significant genes. 



## Set up

```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
              bootstrap.panel = TRUE, 
              bootstrap.show.code = TRUE)
```


```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
library(qvalue)
library(limma)
library(Humanzee)
```


Helper functions
================

Generate regression coefficients and their standard errors.

```{r}
interact2way_coef_voom <- function (eSet, weights) 
{
  fNames <- featureNames(eSet)
  require(nlme)
  gls.res <- lapply(1:length(fNames), function(i) {
    mat1 <- eSet[featureNames(eSet) == fNames[i], ]
    mat1aov.temp <- data.frame(cov = c(t(exprs(mat1))), seqData = mat1$seqData, 
                               species = mat1$species, celline = mat1$celline)
    mat1aov.temp$species <- as.factor(mat1aov.temp$species)
    mat1aov.temp$seqData <- as.factor(mat1aov.temp$seqData)
      fit_null_try <- 
        tryCatch(fit_null <- gls(cov ~ species + seqData, 
                                 weights = weights[i, ], 
                                 data = mat1aov.temp, na.action = na.omit), 
                 condition = function(c) c)
    if (inherits(fit_null_try, "condition")) {
      res <- data.frame(LRatio = NA, LR_pval = NA)
    }
    else {
      fit_interact_try <- 
        tryCatch(fit_interact <- gls(cov ~ species * seqData, 
                                     weights = weights[i, ], 
                                     data = mat1aov.temp, na.action = na.omit), 
                  condition = function(c) c)
    if (inherits(fit_interact_try, "condition")) {
      res <- data.frame(LRatio = NA, LR_pval = NA)
    }
    else {
      aov <- anova(fit_null_try, fit_interact_try)
      res <- data.frame(LRatio = aov[2, "L.Ratio"], 
                        LR_pval = aov[2, "p-value"])
    }
      return(res)
    }
  })
  LR_res <- do.call(rbind, gls.res)
  int.qval <- get_qval(LR_res$LR_pval)
  return(data.frame(ENSGID = fNames, LRatio = LR_res$LRatio, 
                    int.pval = LR_res$LR_pval, int.qval = int.qval))
}
```



Compute voom weights
==========================

`eSetRRP.RP` includes the 3,188 genes that are quantified in all three molecular phenotypes.

```{r, eval = TRUE}
load(file.path(rdadir,"eSetRRP.rda"))
load(file.path(rdadir,"datasub.rda"))

genes_sub <- fData(eSetRRP.RP.Q.log2)$ENSGID

eSet_sub <- eSetNone[ which(fData(eSetNone)$ENSGID %in% genes_sub), 
                      eSetNone$species != "rhesus"]

voom_fit <- 
  Humanzee::voom_NA(counts = exprs(eSet_sub),
                   lib.size = colSums(exprs(eSet_sub), na.rm = TRUE), 
                   design = model.matrix( ~ as.factor(seqData)*as.factor(species), 
                                          data = pData(eSet_sub)), 
                   normalize.method = "quantile")
names(voom_fit)
```


*With voom weights, distribution of the weighted variance.

```{r}
obs_weighted <- voom_fit$E*voom_fit$weights

require(matrixStats)
boxplot(log2(data.frame(human_rna = rowVars(obs_weighted[ , pData(eSet_sub)$species == "human" & pData(eSet_sub)$seqData == "rna"], na.rm = TRUE),
                   human_ribo = rowVars(obs_weighted[ , pData(eSet_sub)$species == "human" & pData(eSet_sub)$seqData == "ribo"], na.rm = TRUE),
                   chimp_rna = rowVars(obs_weighted[ , pData(eSet_sub)$species == "chimp" & pData(eSet_sub)$seqData == "rna"], na.rm = TRUE),
                   chimp_ribo = rowVars(obs_weighted[ , pData(eSet_sub)$species == "chimp" & pData(eSet_sub)$seqData == "ribo"], na.rm = TRUE)
                   ) ) )

```



*Unweighted, distribution of the variances.

```{r}
require(matrixStats)
boxplot(log2(data.frame(human_rna = rowVars(voom_fit$E[ , pData(eSet_sub)$species == "human" & pData(eSet_sub)$seqData == "rna"], na.rm = TRUE),
                   human_ribo = rowVars(voom_fit$E[ , pData(eSet_sub)$species == "human" & pData(eSet_sub)$seqData == "ribo"], na.rm = TRUE),
                   chimp_rna = rowVars(voom_fit$E[ , pData(eSet_sub)$species == "chimp" & pData(eSet_sub)$seqData == "rna"], na.rm = TRUE),
                   chimp_ribo = rowVars(voom_fit$E[ , pData(eSet_sub)$species == "chimp" & pData(eSet_sub)$seqData == "ribo"], na.rm = TRUE)
                   ) ) )

```

t-test
======================

Human vs. Chimp

```{r}
load(file.path(rdadir, "eSetRRP.rda"))

eSet_human_chimp <- eSetRRP.RP.Q.log2[ , 
          eSetRRP.RP.Q.log2$seqData != "protein" & eSetRRP.RP.Q.log2$species != "rhesus"]

fit <- lmFit(voom_fit$E, 
             design = model.matrix( ~ as.factor(seqData)*as.factor(species), 
                                          data = pData(eSet_sub)),
             weights = voom_fit$weights)

contrast_fit <- contrasts.fit(fit, contrasts = cbind(c(0, 0, 0, 1)) )

# Empirical bayes 
contrast_fit <- eBayes(contrast_fit)

# Contrast test of only the interaction effect
contrast_fit$contrast
```


Distribution of the p-values for the interaction effect.

```{r}
hist(contrast_fit$F.p.value, main = "Interaction p-value", breaks = 100)
```


False discovery rate

```{r}
qvals <- qvalue(contrast_fit$F.p.value)$qvalue
sum(qvals < .05)
```





Likelihood ratio test
========================

Human vs. Chimp

```{r}
load(file.path(rdadir, "eSetRRP.rda"))

eSet_human_chimp <- eSetRRP.RP.Q.log2[ , 
          eSetRRP.RP.Q.log2$seqData != "protein" & eSetRRP.RP.Q.log2$species != "rhesus"]

fit_LR_voom <- interact2way_coef_voom(eSet = eSet_human_chimp, 
                                      weights = fit$weights) 

names(fit_LR_voom)
```



Distribution of the p-values for the interaction effect.

```{r}
hist(fit_LR_voom$int.pval, main = "Interaction p-value")
```


False discovery rate

```{r}
qvals <- qvalue(fit_LR_voom$int.pval)$qvalue
sum(qvals < .01)
```



Session information
========================


```{r}
sessionInfo()
```

