---
title: "RNA vs. Ribo divergence, regression coefficients"
author: "Joyce Hsiao"
date: "2015-10-05"
output: 
  html_document:
    toc: true
---

```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```

Compute regression coefficients and compare the ranking with the likelihood ratio statistics.

Set up
======

```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
              bootstrap.panel = TRUE, 
              bootstrap.show.code = TRUE)
```


```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)

library(qvalue)
```



Helper functions
================

Generate regression coefficients and their standard errors.

```{r}
interact2way_coef <- function (eSet) 
{
    fNames <- featureNames(eSet)
    require(nlme)
    gls.res <- lapply(1:length(fNames), function(i) {
        mat1 <- eSet[featureNames(eSet) == fNames[i], ]
        mat1aov.temp <- data.frame(cov = c(t(exprs(mat1))), seqData = mat1$seqData, 
            species = mat1$species, celline = mat1$celline)
        mat1aov.temp$species <- as.factor(mat1aov.temp$species)
        mat1aov.temp$seqData <- as.factor(mat1aov.temp$seqData)
            fit_interact_try <- tryCatch(fit_interact <- gls(cov ~ 
                species * seqData, weights = varIdent(form = ~1 | 
                seqData), data = mat1aov.temp, na.action = na.omit), 
                condition = function(c) c)
                  coef_table <- summary(fit_interact_try)$tTable
                  return(coef_table)
    })
    names(gls.res) <- fNames
    return(gls.res)
}
```



Betas p-value distribution
==========================

Ash works with regression coefficients. We fit the interaction model of two main effects of species and technology and then extract the regression coefficients.

`eSetRRP.RP.Q.log2` includes log2 quantile-normalized RPKM values.

```{r, eval = FALSE}
load(file.path(rdadir,"eSetRRP.rda"))

# Human vs. Chimp
eSet_human_chimp <- eSetRRP.RP.Q.log2[ , 
          eSetRRP.RP.Q.log2$seqData != "protein" & eSetRRP.RP.Q.log2$species != "rhesus"]
riborna_coef_human_chimp <- interact2way_coef(eSet_human_chimp)

# Human vs. Rhesus
eSet_human_rhesus <- eSetRRP.RP.Q.log2[ , 
          eSetRRP.RP.Q.log2$seqData != "protein" & eSetRRP.RP.Q.log2$species != "chimp"]
riborna_coef_human_rhesus <- interact2way_coef(eSet_human_rhesus)

# Chimp vs. Rhesus
eSet_chimp_rhesus <- eSetRRP.RP.Q.log2[ , 
          eSetRRP.RP.Q.log2$seqData != "protein" & eSetRRP.RP.Q.log2$species != "human"]
riborna_coef_chimp_rhesus <- interact2way_coef(eSet_chimp_rhesus)

save(riborna_coef_human_chimp,
     riborna_coef_human_rhesus,
     riborna_coef_chimp_rhesus, file = file.path(rdadir,"riborna_coef.rda"))
```


Distribution of the p-values.

```{r}
load(file = file.path(rdadir, "riborna_coef.rda") )

pvals_human_chimp <- sapply(riborna_coef_human_chimp, function(per_gene) {
                            per_gene[4,4]
                      })
hist(pvals_human_chimp, main = "Human vs. Chimp: p-values", xlab = "p-values")


pvals_human_rhesus <- sapply(riborna_coef_human_rhesus, function(per_gene) {
                            per_gene[4,4]
                      })
hist(pvals_human_rhesus, main = "Human vs. Rhesus: p-values", xlab = "p-values")


pvals_chimp_rhesus <- sapply(riborna_coef_chimp_rhesus, function(per_gene) {
                            per_gene[4,4]
                      })
hist(pvals_chimp_rhesus, main = "Chimp vs. Rhesus: p-values", xlab = "p-values")
```



LR vs. beta p-values
====================

*Human vs. Chimp

```{r}
load(file.path(rdadir, "TEnew.rda"))

plot(x = res.riborna$int.pval,
     y = pvals_human_chimp, xlab = "LR statistic p-values", ylab = "beta p-values",
     xlim = c(0, 1), ylim = c(0, 1))
abline(0, 1, col = "red")
```



*Human vs. Rhesus

```{r}
load(file.path(rdadir, "TE_other_interspecies.rda"))

plot(x = res.riborna_human_rhesus$int.pval,
     y = pvals_human_rhesus, xlab = "LR statistic p-values", ylab = "beta p-values",
     xlim = c(0, 1), ylim = c(0, 1))
abline(0, 1, col = "red")
```



*Chimp vs. Rhesus

```{r}
load(file.path(rdadir, "TE_other_interspecies.rda"))

plot(x = res.riborna_chimp_rhesus$int.pval,
     y = pvals_chimp_rhesus, xlab = "LR statistic p-values", ylab = "beta p-values",
     xlim = c(0, 1), ylim = c(0, 1))
abline(0, 1, col = "red")
```


Venn diagrams
=============

*LRT

```{r}
library(gplots)

kable(data.frame(human_chimp = sum(res.riborna$int.qval < .01),
                 human_rhesus = sum(res.riborna_human_rhesus$int.qval < .01),
                 chimp_rhesus = sum(res.riborna_chimp_rhesus$int.qval < .01) ))

venn(data.frame(human_chimp = res.riborna$int.qval < .01,
                human_rhesus = res.riborna_human_rhesus$int.qval < .01,
                chimp_rhesus = res.riborna_chimp_rhesus$int.qval < .01) )
```



*Betas

```{r}
library(gplots)

qvals_human_chimp <- qvalue(pvals_human_chimp)$qvalue
qvals_human_rhesus <- qvalue(pvals_human_rhesus)$qvalue
qvals_chimp_rhesus <- qvalue(pvals_chimp_rhesus)$qvalue

kable(data.frame(human_chimp = sum(qvals_human_chimp < .01),
                 human_rhesus = sum(qvals_human_rhesus < .01),
                 chimp_rhesus = sum(qvals_chimp_rhesus < .01) ))

venn(data.frame(human_chimp = qvals_human_chimp < .01,
                human_rhesus = qvals_human_rhesus < .01,
                chimp_rhesus = qvals_chimp_rhesus < .01) )
```


Session information
========================


```{r}
sessionInfo()
```

