---
output:
  knitrBootstrap::bootstrap_document:
    title: "RNA-Ribo divergence conditioned on RNA-Protein divergence"
    theme: default
---


RNA-Ribo divergence conditioned on RNA-Protein divergence
===========================================================


```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
               bootstrap.panel = TRUE, bootstrap.show.code = TRUE)
```


```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```


Goal
====

We investigate protein divergence relative to RNA divergence and consider
if buffering at the protein level is likely to be associated with 
buffering at the Ribo level, and if a pattern of non-buffering at the protein level is likely to be associated with a pattern of non-buffering at the
Ribo level.



Definitions
===========

Of all the 3,188 genes included in the analysis, we apply divergence
definition version-20150101 and investigate divergence of genes that
are consistent with the following three patterns. This is the **interaction
approach** described in [Buffering/Version-20150101/Attenuated regulatory
divergence/RNA-protein divergence conditioned on RNA-Ribo divergence](http://jhsiao999.github.io/primate_ribo/project/analysis/TE-genes-ribo-pro-buffering.html).

**Interaction approach**

1. **buffered** at the protein level relative to the RNA level 
    * Sig. different between RNA and protein divergence) *AND* 
      (the absolute value of protein divergence 
      is smaller than the absolute value of RNA divergence) 

2. **non-buffered** at the protein level relative to the RNA level
    * Sig. different between RNA and protein divergence)  *AND*
      (the absolute value of protein divergence is greater 
      than the absolute value of RNA divergence) 
      
3. **null divergence**: of the 3,188 genes, those that do not fall in the 
    above two categories; or equivalently, the genes that do not have a
    pattern that is consistent with significant species difference
    between the RNA and the protein level
    


```{r}
## Load rdas of LRT results and compute fold change estimates using
## RPKM values before normalization. 

## load LRT results of RNA vs. Ribo divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load( file.path(rdadir,"eSetRRP.rda") )

eSet.temp <- eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]

fc.mat <- lapply( seq_along( c("human","chimp") ), function(i) {
  ii <- eSet.temp$species == c("human","chimp")[i]
  eSet.tt <- eSet.temp[, ii]
  emat <- lapply( seq_along( c("ribo","rna","protein") ), function(j) {
    jj <- eSet.tt$seqData == c("ribo","rna","protein")[j]
    rowMeans( exprs(eSet.tt[, jj] ), na.rm=TRUE )
  })
  emat <- do.call(cbind, emat)
  colnames(emat) <- c("ribo","rna","protein")
  return( data.frame(emat) )
})
names(fc.mat) <- c("human","chimp")

dmat_unnormed <- data.frame( ribo = fc.mat$human$ribo-fc.mat$chimp$ribo,
                             rna = fc.mat$human$rna-fc.mat$chimp$rna,
                             pro = fc.mat$human$protein-fc.mat$chimp$protein )

xy.rnapro <- data.frame( rna = dmat_unnormed$rna,
                         pro = dmat_unnormed$pro )
xy.riborna <- data.frame( rna = dmat_unnormed$rna,
                          ribo = dmat_unnormed$ribo )
xy.ribopro <- data.frame( ribo = dmat_unnormed$ribo,
                          pro = dmat_unnormed$pro )
```



## RNA vs. Ribo divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and Ribo
ii_sig_rnaribo <- res.riborna$int.qval < .01

## Genes with RNA > Ribo in effect sizes
ii_rnaGTribo <- abs(xy.riborna$rna) > abs(xy.riborna$ribo)

## Genes with RNA < Ribo in effect sizes
ii_riboGTrna <- abs(xy.riborna$ribo) > abs(xy.riborna$rna)
```


## RNA vs. Protein divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and protein
ii_sig_rnapro <- res.rnapro$int.qval < .01

## Genes with RNA > Protein in effect sizes
ii_rnaGTpro <- abs(xy.rnapro$rna) > abs(xy.rnapro$pro)

## Genes with RNA < Protein in effect sizes
ii_proGTrna <- abs(xy.rnapro$pro) > abs(xy.rnapro$rna)
```


Gene classification
===================

```{r}
buffer_mat <- rbind.data.frame(RNA_GT_Protein = sum(ii_rnaGTpro & ii_sig_rnapro),
                               RNA_LT_Protein = sum(ii_proGTrna & ii_sig_rnapro),
                               RNA_NULL_Protein = sum(!ii_sig_rnapro))
colnames(buffer_mat) <- "# of genes"
kable(buffer_mat)

```





RNA ~ Protein
==============


Among ``r sum( !ii_sig_rnapro )`` genes that are neither 
enhanced nor attenuated at the protein level compared to the RNA level 
(RNA ~ Protein), we identified ``r sum( !ii_sig_rnapro & ii_sig_rnaribo)`` genes 
with significant RNA-Ribo interaction and further classifed these genes as 
follows

* ``r sum( ii_rnaGTribo & ii_sig_rnaribo & !ii_sig_rnapro )`` genes are 
  buffered at the translation level relative to the transcript level, and

* ``r sum( ii_riboGTrna & ii_sig_rnaribo & !ii_sig_rnapro )`` genes are consistent with a pattern of non-buffered genes at the tranlsation level relative to the
  transcript level.



```{r}
ii_null_rnapro <- !(ii_sig_rnapro)
ii_null_rnapro_ribo_buff <- ii_rnaGTribo & ii_sig_rnaribo & !ii_sig_rnapro
ii_null_rnapro_ribo_nonbuff <- ii_riboGTrna & ii_sig_rnaribo & !ii_sig_rnapro

boxplot(abs(dmat_unnormed$rna[ii_null_rnapro]),
        abs(dmat_unnormed$ribo[ii_null_rnapro]),
        abs(dmat_unnormed$pro[ii_null_rnapro]), 
        axes = F, ylim = c(0, 10))
axis(1, at = c(1,2,3), labels = c("RNA", "Ribo", "Protein"), 
     col = "white")
axis(2)
points_cols <- c("Scarlet", "Sunglow", "Shamrock")
points(rep(1, sum(ii_null_rnapro)), 
       abs(dmat_unnormed$rna[ii_null_rnapro]), 
       pch = 19, cex = .5, 
       col = crayon[points_cols[1]] )
points(rep(2, sum(ii_null_rnapro)), 
       abs(dmat_unnormed$ribo[ii_null_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[2]] )
points(rep(3, sum(ii_null_rnapro)), 
       abs(dmat_unnormed$pro[ii_null_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[3]] )
text(2, 10, adj = 0,
     label = paste("RNA ~ Protein", sum(ii_null_rnapro), "genes" ) )
text(2, 9, adj = 0,
     label = paste("RNA ~ Protein & sig. RNA-Ribo", 
                   sum(ii_null_rnapro & ii_sig_rnaribo), "genes" ) )
text(2, 8, adj = 0,
     label = paste("Buffered at translation", 
       sum(ii_null_rnapro_ribo_buff), "genes; \n",
       round(100*sum(ii_null_rnapro_ribo_buff)/
               sum(ii_null_rnapro & ii_sig_rnaribo), 2), "%") )
text(2, 7, adj = 0,
     label = paste("Non-buffered translation", 
       sum(ii_null_rnapro_ribo_nonbuff), "genes; \n",
       round(100*sum(ii_null_rnapro_ribo_nonbuff)/
               sum(ii_null_rnapro & ii_sig_rnaribo), 2), "%") )
title(main = "RNA ~ Protein, no sig. diff. divergence",
      ylab = "absolute divergence (human-chimpanzee fold change)")


# # Compute pairwise p-values between effect sizes of Ribo, RNA, and protein
# # among genes with RNA ~ Protein
# rna_ribo_t <- t.test( abs(dmat_unnormed$rna[ii_proNULLrna]),
#        abs(dmat_unnormed$ribo[ii_proNULLrna]), paired = TRUE )
# rna_pro_t <- t.test( abs(dmat_unnormed$rna[ii_proNULLrna]),
#        abs(dmat_unnormed$pro[ii_proNULLrna]), paired = TRUE )
# ribo_pro_t <- t.test( abs(dmat_unnormed$ribo[ii_proNULLrna]),
#        abs(dmat_unnormed$pro[ii_proNULLrna]), paired = TRUE )
# 
# text(.5, 10, adj = 0,
#      label = paste("Paired t-test of effect sizes") )
# text(.5, 9, adj = 0,
#      label = paste("RNA vs. Ribo: p-value < 1e-14") )
# text(.5, 8, adj = 0,
#      label = paste("RNA vs. Protein: p-value < 1e-15") )
# text(.5, 7, adj = 0,
#      label = paste("Ribo vs. Protein: p-value < 1e-15") )
# 
```


## Signal in RNA ~ Protein


```{r}
## RNA-Ribo buffering or non-buffering
ii_null_rnapro_ribo_buff <- !ii_sig_rnapro & ii_rnaGTribo & ii_sig_rnaribo
ii_null_rnapro_ribo_nonbuff <- !ii_sig_rnapro & ii_riboGTrna & ii_sig_rnaribo

## RNA-Protein buffering or non-buffering
ii_null_rnapro_pro_buff <- !ii_sig_rnapro & ii_rnaGTpro & ii_sig_rnaribo
ii_null_rnapro_pro_nonbuff <- !ii_sig_rnapro & ii_proGTrna & ii_sig_rnaribo

plot_mat <- matrix(0, 2, 2)
rownames(plot_mat) <- c("RNA-Ribo buffered", "RNA-Ribo non-buffered")
colnames(plot_mat) <- c("RNA-Protein buffered", "RNA-Protein non-buffered")

plot_mat[1,1] <- sum(ii_null_rnapro_ribo_buff & ii_null_rnapro_pro_buff)
plot_mat[1,2] <- sum(ii_null_rnapro_ribo_buff & ii_null_rnapro_pro_nonbuff)
plot_mat[2,1] <- sum(ii_null_rnapro_ribo_nonbuff & ii_null_rnapro_pro_buff)
plot_mat[2,2] <- sum(ii_null_rnapro_ribo_nonbuff & ii_null_rnapro_pro_nonbuff)

kable(plot_mat)
```


### Buffered at translation

Among the ``r sum(ii_null_rnapro_ribo_buff)`` genes that were buffered
at the translation level relative to the transcript level, we tested the null hypothesis that these genes are 
as likely to be buffered as to be non-buffered at the protein level relative to 
the transcript level (RNA-Protein). If this null hypothesis were rejected and 
the number of genes that were buffered at the protein layer is larger than 
the number of non-buffered genes at the protein layer, then we say that 
buffering at the translation level is more likey to co-occur with
buffering at the protein level than with non-buffering at the protein level.


```{r}
prop.test(plot_mat[1,1], rowSums(plot_mat)[1])
```


### Not buffered at translation

Among the ``r sum(ii_null_rnapro_ribo_nonbuff)`` genes that were not buffered
at translation, we tested the null hypothesis that these genes are as likely
to be non-buffered as to be buffered at the protein layer. If this null 
hypothesis were rejected and the number of genes that were not buffered at the 
protein layer is more than those buffered at the protein layer, then we 
say that genes that are non-buffered at the translation level are more likely 
to be non-buffered at the protein level than to be buffered at the protein level.


```{r}
## One-proportion z-test
prop.test(plot_mat[2,2], rowSums(plot_mat)[2])

## Exact test of proportions (binomial test)
binom.test(plot_mat[2,2], rowSums(plot_mat)[2])
```





RNA > Protein
==============

Among the genes that are buffered at the protein level compared to the 
RNA level (RNA > protein), we are interested in the number of genes with
buffered translation. In other words, we ask whether genes that were
buffered at the protein level were more likely to be enriched with
genes buffered at the translation level.

For the hypothesis testing, we compute the expected proportion of genes
enriched at the translation level among the genes that were buffered at the
protein level in the set of genes where there is no significant interaction
between RNA and protein (``r sum(ii_null_rnapro)`` genes).


```{r}
boxplot(abs(dmat_unnormed$rna[ii_rnaGTpro & ii_sig_rnapro]),
        abs(dmat_unnormed$ribo[ii_rnaGTpro & ii_sig_rnapro]),
        abs(dmat_unnormed$pro[ii_rnaGTpro & ii_sig_rnapro]), 
        axes = F, ylim = c(0, 10))
axis(1, at = c(1,2,3), labels = c("RNA", "Ribo", "Protein"), 
     col = "white")
axis(2)
points_cols <- c("Scarlet", "Sunglow", "Shamrock")
points(rep(1, sum(ii_rnaGTpro & ii_sig_rnapro)), 
       abs(dmat_unnormed$rna[ii_rnaGTpro & ii_sig_rnapro]), 
       pch = 19, cex = .5, 
       col = crayon[points_cols[1]] )
points(rep(2, sum(ii_rnaGTpro & ii_sig_rnapro)), 
       abs(dmat_unnormed$ribo[ii_rnaGTpro & ii_sig_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[2]] )
points(rep(3, sum(ii_rnaGTpro & ii_sig_rnapro)), 
       abs(dmat_unnormed$pro[ii_rnaGTpro & ii_sig_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[3]] )
text(2.2, 9, adj = 0,
     label = paste("abs(RNA) > abs(Ribo)", 
      sum( ii_rnaGTpro & ii_sig_rnapro & ii_rnaGTribo ), "genes; \n",
      round( 100*sum(ii_rnaGTpro & ii_sig_rnapro & ii_rnaGTribo)/
             sum(ii_rnaGTpro & ii_sig_rnapro), 2),"%") )
title(main = paste("RNA > Protein",sum(ii_rnaGTpro & ii_sig_rnapro),
                   "genes (buffered)"),
      ylab = "absolute divergence (human-chimpanzee fold change)")


## Null proportion of RNA-Protein attenuated | RNA-Ribo attenuated
rnapro_null_pro_buff <- plot_mat[1, 1]/rowSums(plot_mat)[1]


prop.test( sum(ii_rnaGTpro & ii_sig_rnapro & ii_rnaGTribo),
           sum(ii_rnaGTpro & ii_sig_rnapro), p = rnapro_null_pro_buff)

binom.test( sum(ii_rnaGTpro & ii_sig_rnapro & ii_rnaGTribo),
           sum(ii_rnaGTpro & ii_sig_rnapro), p = rnapro_null_pro_buff)
```





RNA < Protein
==============

Among the genes that are not buffered at the protein level compared to the 
RNA level (RNA > protein), we are interested in the number of genes that 
are also not buffered at the translation level.
In other words, we ask whether genes that were
buffered at the protein level were more likely to be enriched with
genes that are also not buffered at the translation level.

For the hypothesis testing, we compute the expected proportion of genes
enriched at the translation level among the genes that were not buffered at the
protein level in the set of genes where there is no significant interaction
between RNA and protein (``r sum(ii_null_rnapro)`` genes).



```{r}
boxplot(abs(dmat_unnormed$rna[ii_proGTrna & ii_sig_rnapro]),
        abs(dmat_unnormed$ribo[ii_proGTrna & ii_sig_rnapro]),
        abs(dmat_unnormed$pro[ii_proGTrna & ii_sig_rnapro]), 
        axes = F, ylim = c(0, 10))
axis(1, at = c(1,2,3), labels = c("RNA", "Ribo", "Protein"), 
     col = "white")
axis(2)
points_cols <- c("Scarlet", "Sunglow", "Shamrock")
points(rep(1, sum(ii_proGTrna & ii_sig_rnapro)), 
       abs(dmat_unnormed$rna[ii_proGTrna & ii_sig_rnapro]), 
       pch = 19, cex = .5, 
       col = crayon[points_cols[1]] )
points(rep(2, sum(ii_proGTrna & ii_sig_rnapro)), 
       abs(dmat_unnormed$ribo[ii_proGTrna & ii_sig_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[2]] )
points(rep(3, sum(ii_proGTrna & ii_sig_rnapro)), 
       abs(dmat_unnormed$pro[ii_proGTrna & ii_sig_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[3]] )
text(1.5, 9, adj = 0,
     label = paste("Protein > RNA", 
                   sum(ii_proGTrna & ii_sig_rnapro), "genes" ) )
text(1.5, 8, adj = 0,
     label = paste("Not buffered at translation", 
      sum(ii_proGTrna & ii_sig_rnapro & ii_riboGTrna), "genes; \n",
      round(100*sum(ii_proGTrna & ii_sig_rnapro & ii_riboGTrna)/
              sum(ii_proGTrna & ii_sig_rnapro),
            2),"%") )
title(main = paste("RNA < Protein",sum(ii_proGTrna & ii_sig_rnapro),
                  "genes (Non-buffering)"),
      ylab = "absolute divergence (human-chimpanzee fold change)")


## Null proportion of RNA-Ribo enhanced | RNA-Protein enhanced
rnapro_null_pro_nonbuff <- plot_mat[2,2]/rowSums(plot_mat)[2]

prop.test( sum(ii_proGTrna & ii_sig_rnapro & ii_riboGTrna),
           sum(ii_proGTrna & ii_sig_rnapro), p = rnapro_null_pro_nonbuff)

binom.test( sum(ii_proGTrna & ii_sig_rnapro & ii_riboGTrna),
           sum(ii_proGTrna & ii_sig_rnapro), p = rnapro_null_pro_nonbuff)
```


Session information
===================

```{r}
sessionInfo()
```

