---
output:
  knitrBootstrap::bootstrap_document:
    title: "RNA-Ribo divergence conditioned on RNA-Protein divergence"
    theme: default
---


RNA-Ribo divergence conditioned on RNA-Protein divergence
===========================================================


```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
               bootstrap.panel = TRUE, bootstrap.show.code = TRUE)
```


```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```


Goal
====

Here we consider a set of genes with sig. RNA-protein interaction and 
investigate the contribution of buffering at the translation level
to the protein level. Two approaches are discussed


Gene classification
===================

```{r}
## Load rdas of LRT results and compute fold change estimates using
## RPKM values before normalization. 

## load LRT results of RNA vs. Ribo divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load( file.path(rdadir,"eSetRRP.rda") )

eSet.temp <- eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]

fc.mat <- lapply( seq_along( c("human","chimp") ), function(i) {
  ii <- eSet.temp$species == c("human","chimp")[i]
  eSet.tt <- eSet.temp[, ii]
  emat <- lapply( seq_along( c("ribo","rna","protein") ), function(j) {
    jj <- eSet.tt$seqData == c("ribo","rna","protein")[j]
    rowMeans( exprs(eSet.tt[, jj] ), na.rm=TRUE )
  })
  emat <- do.call(cbind, emat)
  colnames(emat) <- c("ribo","rna","protein")
  return( data.frame(emat) )
})
names(fc.mat) <- c("human","chimp")

dmat_unnormed <- data.frame( ribo = fc.mat$human$ribo-fc.mat$chimp$ribo,
                             rna = fc.mat$human$rna-fc.mat$chimp$rna,
                             pro = fc.mat$human$protein-fc.mat$chimp$protein )

xy.rnapro <- data.frame( rna = dmat_unnormed$rna,
                         pro = dmat_unnormed$pro )
xy.riborna <- data.frame( rna = dmat_unnormed$rna,
                          ribo = dmat_unnormed$ribo )
xy.ribopro <- data.frame( ribo = dmat_unnormed$ribo,
                          pro = dmat_unnormed$pro )
```



## RNA vs. Ribo divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and Ribo
ii_sig_rnaribo <- res.riborna$int.qval < .01

## Genes with RNA > Ribo in effect sizes
ii_rnaGTribo <- abs(xy.riborna$rna) > abs(xy.riborna$ribo)

## Genes with RNA < Ribo in effect sizes
ii_riboGTrna <- abs(xy.riborna$ribo) > abs(xy.riborna$rna)
```


## RNA vs. Protein divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and protein
ii_sig_rnapro <- res.rnapro$int.qval < .01

## Genes with RNA > Protein in effect sizes
ii_rnaGTpro <- abs(xy.rnapro$rna) > abs(xy.rnapro$pro)

## Genes with RNA < Protein in effect sizes
ii_proGTrna <- abs(xy.rnapro$pro) > abs(xy.rnapro$rna)
```



Interaction approach
====================

**Interaction appoach**: take the set of genes with sig. RNA-protein interaction, 
select a subset of genes with sig. RNA-ribo interaction. Then, we divided the
genes into two categories

    * Buffered at the protein level: abs(RNA fold change) > abs(Protein fold change)
    * Non-Buffered at the protein level: abs(RNA fold change) < abs(Protein fold change)


```{r}
interact_mat <- 
    matrix(0, 3, 4, dimnames = list(c("Protein bufferd (& sig. RNA-protein)", 
                                  "Protein non-buffered (& sig. RNA-protein)",
                                  "No sig. RNA-Protein interaction"),
                                c("Ribo buffered (& sig. RNA-Ribo)",
                                  "Ribo non-buffered (& sig. RNA-Ribo)",
                                  "No sig. RNA-Ribo interaction",
                                  "Row sum")))

interact_mat[1,4] <- sum(ii_sig_rnapro & ii_rnaGTpro)
interact_mat[2,4] <- sum(ii_sig_rnapro & ii_proGTrna)
interact_mat[3,4] <- sum(!ii_sig_rnapro)

interact_mat[1,1] <- sum(ii_sig_rnapro & ii_rnaGTpro &
                         ii_sig_rnaribo & ii_rnaGTribo)
interact_mat[1,2] <- sum(ii_sig_rnapro & ii_rnaGTpro &
                         ii_sig_rnaribo & ii_riboGTrna)
interact_mat[1,3] <- sum(ii_sig_rnapro & ii_rnaGTpro &
                         !ii_sig_rnaribo)

interact_mat[2,1] <- sum(ii_sig_rnapro & ii_proGTrna &
                           ii_sig_rnaribo & ii_rnaGTribo)
interact_mat[2,2] <- sum(ii_sig_rnapro & ii_proGTrna &
                           ii_sig_rnaribo & ii_riboGTrna)
interact_mat[2,3] <- sum(ii_sig_rnapro & ii_proGTrna &
                           !ii_sig_rnaribo)

interact_mat[3,1] <- sum(!ii_sig_rnapro & 
                           ii_sig_rnaribo & ii_rnaGTribo)
interact_mat[3,2] <- sum(!ii_sig_rnapro &
                           ii_sig_rnaribo & ii_riboGTrna)
interact_mat[3,3] <- sum(!ii_sig_rnapro & 
                           !ii_sig_rnaribo)

kable(interact_mat)
```



Consider among the ``r sum(ii_sig_rnapro & ii_rnaGTpro & ii_sig_rnaribo)`` genes
with significant RNA-protein interaction that are buffered at the protein level 
and with significant RNA-Ribo interaction, we test the null hypothesis that 
these genes are as likely to be buffered and to be non-buffered at the Ribo 
level.

We will compute the expected number of genes using the set of genes that have
no significant interaction between RNA and Protein.


```{r}
# Compute the number of genes in the null set
ii_null_rnapro_buff <- !ii_sig_rnapro & ii_rnaGTpro & ii_sig_rnaribo

# Compute the number of genes that are buffered at the Ribo level
ii_null_rnapro_buff_ribo <- !ii_sig_rnapro & ii_rnaGTpro & 
                            ii_sig_rnaribo & ii_rnaGTribo
  

prop.test(interact_mat[1,1], interact_mat[1,1] + interact_mat[1,2],
           p = sum(ii_null_rnapro_buff_ribo)/sum(ii_null_rnapro_buff) )

binom.test(interact_mat[1,1], interact_mat[1,1] + interact_mat[1,2],
           p = sum(ii_null_rnapro_buff_ribo)/sum(ii_null_rnapro_buff) )
```



Size approach
====================

**Size approach**: take the set of genes with sig. RNA-protein interaction. Then, we divided the genes into two categories

    * Buffered at the protein level: abs(RNA fold change) > abs(Protein fold change)
    * Non-Buffered at the protein level: abs(RNA fold change) < abs(Protein fold change)


```{r}
size_mat <- 
    matrix(0, 3, 3, dimnames = list(c("Protein bufferd (& sig. RNA-protein)", 
                                  "Protein non-buffered (& sig. RNA-protein)",
                                  "No sig. RNA-Protein interaction"),
                                c("Ribo buffered (& sig. RNA-Ribo)",
                                  "Ribo non-buffered (& sig. RNA-Ribo)",
                                  "Row sum")))

size_mat[1,1] <- sum(ii_sig_rnapro & ii_rnaGTpro &
                         ii_rnaGTribo)
size_mat[1,2] <- sum(ii_sig_rnapro & ii_rnaGTpro &
                         ii_riboGTrna)
size_mat[1,3] <- sum(ii_sig_rnapro & ii_rnaGTpro)

size_mat[2,1] <- sum(ii_sig_rnapro & ii_proGTrna &
                           ii_rnaGTribo)
size_mat[2,2] <- sum(ii_sig_rnapro & ii_proGTrna &
                           ii_riboGTrna)
size_mat[2,3] <- sum(ii_sig_rnapro & ii_proGTrna)

size_mat[3,1] <- sum(!ii_sig_rnapro & 
                           ii_rnaGTribo)
size_mat[3,2] <- sum(!ii_sig_rnapro &
                           ii_riboGTrna)
size_mat[3,3] <- sum(!ii_sig_rnapro)

kable(size_mat)
```



Consider among the ``r sum(ii_sig_rnapro & ii_rnaGTpro)`` genes
with significant RNA-protein interaction that are buffered at the protein level, 
we test the null hypothesis that 
these genes are as likely to be buffered and to be non-buffered at the Ribo 
level.

We will compute the expected number of genes using the set of genes that have
no significant interaction between RNA and Protein.


```{r}
# Compute the number of genes in the null set
ii_null_rnapro_buff_size <- !ii_sig_rnapro & ii_rnaGTpro

# Compute the number of genes that are buffered at the Ribo level
ii_null_rnapro_buff_size_ribo <- !ii_sig_rnapro & ii_rnaGTpro &
                                   ii_rnaGTribo
  

prop.test(size_mat[1,1], size_mat[1,1] + size_mat[1,2],
           p = sum(ii_null_rnapro_buff_size_ribo)/
               sum(ii_null_rnapro_buff_size) )

binom.test(size_mat[1,1], size_mat[1,1] + size_mat[1,2],
           p = sum(ii_null_rnapro_buff_size_ribo)/
               sum(ii_null_rnapro_buff_size) )
```






Session information
===================

```{r}
sessionInfo()
```

