---
output:
  knitrBootstrap::bootstrap_document:
    title: "RNA vs. Protein divergence"
    theme: default
    theme.chooser: TRUE
    highlight.chooser: TRUE
---

RNA vs. Protein divergence
=================================

```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


[view source](compare-rna-pro-divergence.Rmd)


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
              bootstrap.panel = TRUE, 
              bootstrap.show.code = TRUE)
```


Compute the number of genes attenuated or enhanced at the protein level 
compared to the RNA level.


Definitions
============

The set of genes that are significant in inter-species difference of 
RNA vs. protein can be divided into two categories, and within each of the categories, the genes can further be
divided into two subcategories as follows

1. No sig. inter-species difference in RNA

    1.1. No sig. inter-species difference in Protein
    
    1.2. Sig. inter-species difference in Protein
    
2. Sig. inter-species difference in RNA

    2.1. ( abs(inter-species RNA FC) > abs(inter-species Protein FC) ) OR
         ( inter-species RNA FC and inter-species Protein FC are of 
           the opposite direction) - **Attenuation at the protein level**
           
    2.2. ( abs(inter-species RNA FC) < abs(inter-species Protein FC) ) AND
         ( inter-species RNA FC and inter-species Protein FC are of the
           same direction) - **Enhancement at the protein level**



```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```



 
Load rdas of LRT results and compute fold change estimates using
RPKM values before quantile normalization. 


```{r, bootstrap.show.code = TRUE}
## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))

## load LRT results of Ribo vs. RNA divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of DE analysis
load(file.path(rdadir,"DE.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load(file.path(rdadir,"eSetRRP.rda"))
eSet.temp = eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]
fc.mat=lapply(seq_along(c("human","chimp")), function(i) {
  ii = eSet.temp$species==c("human","chimp")[i]
  eSet.tt = eSet.temp[,ii]
  emat = lapply(seq_along(c("ribo","rna","protein")),function(j) {
    jj = eSet.tt$seqData==c("ribo","rna","protein")[j]
    rowMeans(exprs(eSet.tt[,jj]),na.rm=TRUE)
  })
  emat=do.call(cbind,emat)
  colnames(emat)=c("ribo","rna","protein")
  return(data.frame(emat))
})
names(fc.mat)=c("human","chimp")
dmat_unnormed = data.frame(ribo=fc.mat$human$ribo-fc.mat$chimp$ribo,
                           rna=fc.mat$human$rna-fc.mat$chimp$rna,
                           pro=fc.mat$human$protein-fc.mat$chimp$protein)

xy.rnapro = data.frame(rna=dmat_unnormed$rna,
                       pro=dmat_unnormed$pro)
xy.riborna = data.frame(rna=dmat_unnormed$rna,
                        ribo=dmat_unnormed$ribo)
xy.ribopro = data.frame(ribo=dmat_unnormed$ribo,
                        pro=dmat_unnormed$pro)
```


Define subsets of genes at FDR < .01.

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and protein
ii_sig_rnapro <- res.rnapro$int.qval < .01

## Genes with sig. human/chimp difference in RNA levels
ii_sig_RNA_DE <- rnaRes.Q$qval < .01

## Genes with sig. human/chimp difference in Protein levels
ii_sig_Pro_DE <- massRes.Q$qval < .01


## Category 1 - Enhanced divergence, criteria are as follows
## a) Sig. RNA vs. Protein difference, AND
## b) Sig. DE at RNA level, AND
## c) ( abs(RNA FC) < abs(Protein FC) ) AND ( RNA FC * Protein FC > 0)
ii_proGTrna <- 
    (ii_sig_rnapro) & ii_sig_RNA_DE & 
    ( (abs(xy.rnapro$pro) > abs(xy.rnapro$rna)) & 
      (xy.rnapro$rna * xy.rnapro$pro > 0 ) )


## Category 2 - Attenuated divergence, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) Sig. DE at RNA level, AND 
## c) ( abs(RNA FC) > abs(Protein FC) ) AND ( RNA FC * Protein FC < 0)
ii_rnaGTpro <- 
    (ii_sig_rnapro) & ii_sig_RNA_DE & 
    ( (abs(xy.rnapro$rna) > abs(xy.rnapro$pro) ) | 
      (xy.rnapro$rna * xy.rnapro$pro < 0) ) 


## Category 3, Pseudo enhacement, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) No sig. DE at RNA level, AND 
## c) Sig. DE at Protein level
ii_null_rnaGTpro <- 
    ii_sig_rnapro & !ii_sig_RNA_DE & ii_sig_Pro_DE


## Category 4, Null divergence, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) No sig. DE at RNA level, AND 
## c) No sig. DE at Protein level
ii_null_rnaNULLprotein <- 
    ii_sig_rnapro & !ii_sig_RNA_DE & !ii_sig_Pro_DE


Protein_refmat <- rbind.data.frame(attenuated = sum(ii_rnaGTpro),
                                   enhanced = sum(ii_proGTrna),
                                   null_enhanced = sum(ii_null_rnaGTpro),
                                   null = sum(ii_null_rnaNULLprotein) )
colnames(Protein_refmat) <- "# genes"
Protein_refmat
```






Session information
===========

```{r}
sessionInfo()
```

