---
output:
  knitrBootstrap::bootstrap_document:
    title: "RNA vs. Ribo divergence"
    theme: default
    theme.chooser: TRUE
    highlight.chooser: TRUE
---

RNA vs. Ribo divergence
=================================

```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


[view source](compare-ribo-rna-direction.Rmd)


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
              bootstrap.panel = TRUE, 
              bootstrap.show.code = TRUE)
```


Compute the number of genes attenuated or enhanced at the Ribo level 
compared to the RNA level.


Definitions
============

The set of genes that are significant in inter-species difference of 
RNA vs. Ribo (translation efficiency) can be divided into two categories, and within each of the categories, the genes can further be divided into two subcategories as follows

1. No sig. inter-species difference in RNA

    1.1. No sig. inter-species difference in Ribo
    
    1.2. Sig. inter-species difference in Ribo
    
2. Sig. inter-species difference in RNA

    2.1. ( abs(inter-species RNA FC) > abs(inter-species Ribo FC) ) OR
         ( inter-species RNA FC and inter-species Ribo FC are of 
           the opposite direction) - **Attenuation at the Ribo level**
           
    2.2. ( abs(inter-species RNA FC) < abs(inter-species Ribo FC) ) AND
         ( inter-species RNA FC and inter-species Ribo FC are of the
           same direction) - **Enhancement at the Ribo level**



```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```



 
Load rdas of LRT results and compute fold change estimates using
RPKM values before quantile normalization. 


```{r, bootstrap.show.code = TRUE}
## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))

## load LRT results of Ribo vs. RNA divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of DE analysis
load(file.path(rdadir,"DE.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load(file.path(rdadir,"eSetRRP.rda"))
eSet.temp = eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]
fc.mat=lapply(seq_along(c("human","chimp")), function(i) {
  ii = eSet.temp$species==c("human","chimp")[i]
  eSet.tt = eSet.temp[,ii]
  emat = lapply(seq_along(c("ribo","rna","protein")),function(j) {
    jj = eSet.tt$seqData==c("ribo","rna","protein")[j]
    rowMeans(exprs(eSet.tt[,jj]),na.rm=TRUE)
  })
  emat=do.call(cbind,emat)
  colnames(emat)=c("ribo","rna","protein")
  return(data.frame(emat))
})
names(fc.mat)=c("human","chimp")
dmat_unnormed = data.frame(ribo=fc.mat$human$ribo-fc.mat$chimp$ribo,
                           rna=fc.mat$human$rna-fc.mat$chimp$rna,
                           pro=fc.mat$human$protein-fc.mat$chimp$protein)

xy.rnapro = data.frame(rna=dmat_unnormed$rna,
                       pro=dmat_unnormed$pro)
xy.riborna = data.frame(rna=dmat_unnormed$rna,
                        ribo=dmat_unnormed$ribo)
xy.ribopro = data.frame(ribo=dmat_unnormed$ribo,
                        pro=dmat_unnormed$pro)
```


Define subsets of genes at FDR < .01.

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and Ribo
ii_sig_rnaribo <- res.riborna$int.qval < .01

## Genes with sig. human/chimp difference in RNA levels
ii_sig_RNA_DE <- rnaRes.Q$qval < .01

## Genes with sig. human/chimp difference in Ribo levels
ii_sig_Ribo_DE <- riboRes.Q$qval < .01


## Category 1 - Enhanced divergence, criteria are as follows
## a) Sig. RNA vs. Ribo difference, AND
## b) Sig. DE at RNA level, AND
## c) ( abs(RNA FC) < abs(Ribo FC) ) AND ( RNA FC * Ribo FC > 0)
ii_riboGTrna <- 
    (ii_sig_rnaribo) & ii_sig_RNA_DE & 
    ( (abs(xy.riborna$ribo) > abs(xy.riborna$rna)) & 
      (xy.riborna$rna * xy.riborna$ribo > 0 ) )


## Category 2 - Attenuated divergence, criteria are as follows
## a) Sig. RNA vs. Ribo diff., AND
## b) Sig. DE at RNA level, AND 
## c) ( abs(RNA FC) > abs(Ribo FC) ) AND ( RNA FC * Ribo FC < 0)
ii_rnaGTribo <- 
    (ii_sig_rnaribo) & ii_sig_RNA_DE & 
    ( (abs(xy.riborna$rna) > abs(xy.riborna$ribo) ) | 
      (xy.riborna$rna * xy.riborna$ribo < 0) ) 


## Category 3, Pseudo enhacement, criteria are as follows
## a) Sig. RNA vs. Ribo diff., AND
## b) No sig. DE at RNA level, AND 
## c) Sig. DE at Ribo level
ii_null_rnaGTribo <- 
    ii_sig_rnaribo & !ii_sig_RNA_DE & ii_sig_Ribo_DE


## Category 4, Null divergence, criteria are as follows
## a) Sig. RNA vs. Ribo diff., AND
## b) No sig. DE at RNA level, AND 
## c) No sig. DE at Ribo level
ii_null_rnaNULLribo <- 
    ii_sig_rnaribo & !ii_sig_RNA_DE & !ii_sig_Ribo_DE


Ribo_refmat <- rbind.data.frame(attenuated = sum(ii_rnaGTribo),
                                   enhanced = sum(ii_riboGTrna),
                                   null_enhanced = sum(ii_null_rnaGTribo),
                                   null = sum(ii_null_rnaNULLribo) )
colnames(Ribo_refmat) <- "# genes"
Ribo_refmat
```






Session information
===========

```{r}
sessionInfo()
```

