---
output:
  knitrBootstrap::bootstrap_document:
    title: "Compare transcript divergence with translation divergence - 
            including genes that may not be quantified in Mass Spec data "
    theme: default
    theme.chooser: TRUE
    highlight.chooser: TRUE
---


Differential divergence analysis
=================================

```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


[view source](compare-ribo-rna-quantified-in-sequencing.Rmd)


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
              bootstrap.panel = TRUE, 
              bootstrap.show.code = TRUE)
```



```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```



Goal
====

We analyzed all three pairwise species comparisons of translation efficiency
using a set of genes that may not be quantified in Mass Spectrometry data. 
This step is to investigate and ascertain results obtained from a smaller set 
of genes that are quantified across all three phenotypes.



Inter-species comparisons
=========================

We used a likelihood ratio test to compare species divergence 
at the Ribo level versus at the RNA level.


```{r, eval = F}

# Load rdas of genes quantified in RNA and Ribo
# eSetRPKM: RPKM values 
# eSetGM: RPKM values relative to the standard cell line
# eSetRPKM.Q and eSetGM.Q: data quantile-normalized within each data type
load(file.path(rdadir,"normeddatasub.rda"))


# log transform the RPKM values
emat <- exprs(eSetRPKM.Q)
emat.log2 <- log2(emat)
eSetRPKM.Q.log2 <- ExpressionSet(assayData = emat.log2,
                            phenoData = phenoData(eSetRPKM.Q),
                            experimentData = experimentData(eSetRPKM.Q) )
featureData(eSetRPKM.Q.log2) <- featureData(eSetRPKM.Q)                  


# Compare human and chimp
eset_human_chimp <- eSetRPKM.Q.log2[ ,eSetRPKM.Q.log2$species != "rhesus"]
res.riborna_full_human_chimp <- interact2way(eset_human_chimp)

# Compare chimp and rhesus
eset_chimp_rhesus <- eSetRPKM.Q.log2[ ,eSetRPKM.Q.log2$species != "human"]
res.riborna_full_chimp_rhesus <- interact2way(eset_chimp_rhesus)

# Compare human and rhesus
eset_human_rhesus <- eSetRPKM.Q.log2[ ,eSetRPKM.Q.log2$species != "chimp"]
res.riborna_full_human_rhesus <- interact2way(eset_human_rhesus)

save(eSetRPKM.Q.log2,
     res.riborna_full_human_chimp,
     res.riborna_full_chimp_rhesus,
     res.riborna_full_human_rhesus, 
     file = file.path(rdadir,"TE_full.rda"))
```


The genes with significant interaction effect at FDR < .01 are considered
as significant in species difference between RNA and Ribo, or equivalently,
species difference in translation efficiency. 


```{r}
load(file.path(rdadir, "TE_full.rda"))

ngenes <- dim(res.riborna_full_human_chimp)[1]

# Check the number of genes for which
# the likelihood ratio test statistic can not be computed

na_mat <- data.frame(total_genes = ngenes,
                     human_chimp_na = sum(is.na(res.riborna_full_human_chimp$int.pval)),
                     human_rhesus_na = sum(is.na(res.riborna_full_human_rhesus$int.pval)),
                     chimp_rhesus_na = sum(is.na(res.riborna_full_chimp_rhesus$int.pval)) )
kable(na_mat)
```


Exclude the two NA genes.

```{r}
ii_exclude_genes <- is.na(res.riborna_full_human_rhesus$int.pval)

exprs(eSetRPKM.Q.log2)[which(ii_exclude_genes), ]
```



```{r}
sig_genes <- 
    data.frame(total_genes = ngenes - sum(ii_exclude_genes),
               human_chimp = sum(res.riborna_full_human_chimp$int.qval[
                                which(!ii_exclude_genes)] < .01, na.rm = TRUE),
               human_rhesus = sum(res.riborna_full_human_rhesus$int.qval[
                                which(!ii_exclude_genes)] < .01, na.rm = TRUE),
               chimp_rhesus = sum(res.riborna_full_chimp_rhesus$int.qval[
                                which(!ii_exclude_genes)] < .01, na.rm = TRUE) )
kable(sig_genes)
```



p-value distributions
=====================

```{r}
par(mfrow = c(2,2))
plot_list <- list(human_chimp = res.riborna_full_human_chimp$int.pval,
                  chimp_rhesus = res.riborna_full_chimp_rhesus$int.pval,
                  human_rhesus = res.riborna_full_human_rhesus$int.pval)
for (i in 1: length(plot_list)) {
  hist(plot_list[[i]], xlab = "p-value", main = names(plot_list)[i] )
}
```




q-value distributions
=====================

```{r}
par(mfrow = c(2,2))
plot_list <- list(human_chimp = res.riborna_full_human_chimp$int.qval,
                  chimp_rhesus = res.riborna_full_chimp_rhesus$int.qval,
                  human_rhesus = res.riborna_full_human_rhesus$int.qval)
for (i in 1: length(plot_list)) {
  hist(plot_list[[i]], xlab = "q-value", main = names(plot_list)[i] )
}
```




BH-adjusted p-value distributions
=================================

```{r}
par(mfrow = c(2,2))
plot_list <- list(human_chimp = p.adjust(res.riborna_full_human_chimp$int.pval,
                                         method = "hochberg"),
                  chimp_rhesus = p.adjust(res.riborna_full_chimp_rhesus$int.pval,
                                          method = "hochberg"),
                  human_rhesus = p.adjust(res.riborna_full_human_rhesus$int.pval,
                                          method = "hochberg") )
for (i in 1: length(plot_list)) {
  hist(plot_list[[i]], xlab = "BH adjusted p-value", main = names(plot_list)[i] )
}

# human vs. chimp BH adjusted p-values
summary(plot_list[[1]]) 

# chimp vs. rhesus BH adjusted p-values
summary(plot_list[[2]]) 

# human vs. rhesus
summary(plot_list[[3]]) 
```










Session information
====================

```{r}
sessionInfo()
```

