---
output:
  knitrBootstrap::bootstrap_document:
    title: "Differential expression analysis separatley for each molecular phenotype"
    theme: default
---


Differential expression analysis 
=================================

```{r run_date, results='asis', echo=FALSE}
email = "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
cat(paste("Author:", email, "\n Last updated:", last_update))
```



```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
library(Biobase)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(fig.width = 800/96, fig.height = 800/96, dpi = 96,
              message = FALSE, warning = FALSE, eval = TRUE, echo = TRUE,
              bootstrap.panel = FALSE, bootstrap.show.code = FALSE)
```


```{r setUp}
library(devtools)
install_github("jhsiao999/primate_ribo/Humanzee")
require(Humanzee)


library(Biobase)
rm(list=ls())
dir <- "../../project"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")

# source(file.path( codedir, "prepareDivergencePlots.R"))
```


Definitions/Terminologies
=========================

*Divergence* is a term used in evolutionary studies referring to inter-species difference. We studied inter-primate divergence at the transcript level, translation level, and protein level.


Method
======
A nested linear model framework. Same approach as in Zia's paper. 



Ribosome occupancy
==================

```{r, eval = F}

## Load differential expression results and else
load(file.path(rdadir, "eSetRiboProfile.rda"))

# Human vs. Chimpanzee
sum(ribo.Res.hc$qval < .01)

# Human vs. Rhesus
sum(ribo.Res.hr$qval < .01)

# Chimp vs. Rhesus
sum(ribo.Res.rc$qval < .01)

load(file.path(rdadir, "DE.rda"))

```



Session information
===================

```{r}
sessionInfo()
```


