---
output:
  knitrBootstrap::bootstrap_document:
    title: "Sanity check on RNA vs. Protein divergence results"
    theme: default
    theme.chooser: TRUE
    highlight.chooser: TRUE
---


Checking RNA vs. Protein divergence
====================================

```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


[view source](divergence-check-rna-ref.Rmd)


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
              bootstrap.panel = TRUE, 
              bootstrap.show.code = TRUE)
```



Definitions
============

The set of genes that are significant in inter-species difference of 
RNA vs. Protein divergence can be divided into two categories, and within
each of the categories, the genes can further be divided into two 
subcategories as follows

1. No sig. inter-species difference in Protein

    1.1. No sig. inter-species difference in RNA
    
    1.2. Sig. inter-species difference in RNA
    
2. Sig. inter-species difference in Protein

    2.1. ( abs(inter-species Protein FC) > abs(inter-species RNA FC) ) OR
         ( inter-species Protein FC and inter-species Ribo FC are of 
           the opposite direction) - **Attenuation at the RNA level**
           
    2.2. ( abs(inter-species Protein FC) < abs(inter-species RNA FC) ) AND
         ( inter-species RNA FC and inter-species Protein FC are of the
           same direction) - **Enhancement at the RNA level**



```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```



 
Load rdas of LRT results and compute fold change estimates using
RPKM values before normalization. 


```{r, bootstrap.show.code = TRUE}
## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))

## load LRT results of Ribo vs. RNA divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of DE analysis
load(file.path(rdadir,"DE.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load(file.path(rdadir,"eSetRRP.rda"))
eSet.temp = eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]
fc.mat=lapply(seq_along(c("human","chimp")), function(i) {
  ii = eSet.temp$species==c("human","chimp")[i]
  eSet.tt = eSet.temp[,ii]
  emat = lapply(seq_along(c("ribo","rna","protein")),function(j) {
    jj = eSet.tt$seqData==c("ribo","rna","protein")[j]
    rowMeans(exprs(eSet.tt[,jj]),na.rm=TRUE)
  })
  emat=do.call(cbind,emat)
  colnames(emat)=c("ribo","rna","protein")
  return(data.frame(emat))
})
names(fc.mat)=c("human","chimp")
dmat_unnormed = data.frame(ribo=fc.mat$human$ribo-fc.mat$chimp$ribo,
                           rna=fc.mat$human$rna-fc.mat$chimp$rna,
                           pro=fc.mat$human$protein-fc.mat$chimp$protein)

xy.rnapro = data.frame(rna=dmat_unnormed$rna,
                       pro=dmat_unnormed$pro)
xy.riborna = data.frame(rna=dmat_unnormed$rna,
                        ribo=dmat_unnormed$ribo)
xy.ribopro = data.frame(ribo=dmat_unnormed$ribo,
                        pro=dmat_unnormed$pro)
```


Define subsets of genes at FDR < .01.

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and protein
ii_sig_rnapro <- res.rnapro$int.qval < .01

## Genes with sig. human/chimp difference in RNA levels
ii_sig_RNA_DE <- rnaRes.Q$qval < .01

## Genes with sig. human/chimp difference in Protein levels
ii_sig_Pro_DE <- massRes.Q$qval < .01

## Category 1 - Enhanced divergence, criteria are as follows
## a) Sig. RNA vs. Protein difference, AND
## b) Sig. diff. at Protein level, AND
## c) ( abs(Protein FC) < abs(RNA FC) ) AND ( RNA FC * Protein FC > 0)
iicheck_rnaGTpro <- 
    (ii_sig_rnapro) & ii_sig_Pro_DE & 
    ( (abs(xy.rnapro$rna) > abs(xy.rnapro$pro)) & 
      (xy.rnapro$rna * xy.rnapro$pro > 0 ) )


## Category 2 - Attenuated divergence, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) Sig. DE at Protein level, AND 
## c) ( abs(Protein FC) > abs(RNA FC) ) AND ( RNA FC * Protein FC < 0)
iicheck_proGTrna <- 
    (ii_sig_rnapro) & ii_sig_Pro_DE & 
    ( (abs(xy.rnapro$pro) > abs(xy.rnapro$rna) ) | 
      (xy.rnapro$rna * xy.rnapro$pro < 0) ) 


## Category 3, Pseudo enhacement, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) No sig. DE at Protein level, AND 
## c) Sig. DE at RNA level
iicheck_null_proGTrna <- 
    ii_sig_rnapro & !ii_sig_Pro_DE & ii_sig_RNA_DE


## Category 4, Null divergence, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) No sig. DE at RNA level, AND 
## c) No sig. DE at Protein level
iicheck_null_proNULLrna <- 
    ii_sig_rnapro & !ii_sig_RNA_DE & !ii_sig_Pro_DE


RNA_refmat <- rbind.data.frame(attenuated = sum(iicheck_proGTrna),
                               enhanced = sum(iicheck_rnaGTpro),
                               null_enhanced = sum(iicheck_null_proGTrna),
                               null = sum(iicheck_null_proNULLrna) )
colnames(RNA_refmat) <- "# genes"
RNA_refmat
```






Session information
===========

```{r}
sessionInfo()
```

