---
title: "Simulations"
output: html_document
---

Date: 2015-09-03

Goal
====

We wonder what the expected (null) cell frequencies of this 2 x 2 table: Ribo buffered or non-buffered versus Protein buffered or non-buffered. 

Ribo buffered is defined as the scenario under which a gene's effect size at the RNA level is greater than the effect size at the Ribo level. Effect size is computed as the absolute value of log2 fold change of expression values between Human and Chimpanzee. 


Experiment
==========

For each phenotype, we randomly sample integers from -10 to 10. Negative
numbers represent the case when expression levels in chimp are lower
than expression levels in human. 

```{r}
rna <- c(sample(1:10, size = 100, replace = TRUE),
         -1*sample(1:10, size = 100, replace = TRUE) )

ribo <- c(sample(1:10, size = 100, replace = TRUE),
         -1*sample(1:10, size = 100, replace = TRUE) )

pro <- c(sample(1:10, size = 100, replace = TRUE),
         -1*sample(1:10, size = 100, replace = TRUE) )
data <- cbind(rna, ribo, pro)
str(data)
```

Define significant interaction as when the simulated RNA value here
is the same as the simulated Ribo value, and similarly for RNA 
versus Protein. We exclude the genes that do not have significant
interaction between RNA and Ribo *AND* and also between RNA and protein.


```{r}
ii_rnaribo <- rna != ribo
ii_rnapro <- rna != pro
data_sub <- as.data.frame(data[ii_rnaribo & ii_rnapro, ])
head(data_sub)
```

Construct a buffering type table based on our simulated data.

```{r}
buff_table <- 
    matrix(0, 2, 2,
          dimnames = list(RNA_Ribo = c("buffered", "non-buffered"),
                          RNA_Protein = c("buffered", "non-buffered")))

buff_table[1,1] <- sum(abs(data_sub$rna) > abs(data_sub$ribo) &
                       abs(data_sub$rna) > abs(data_sub$pro))
buff_table[1,2] <- sum(abs(data_sub$rna) > abs(data_sub$ribo) &
                       abs(data_sub$pro) > abs(data_sub$rna))
buff_table[2,1] <- sum(abs(data_sub$ribo) > abs(data_sub$rna) &
                       abs(data_sub$rna) > abs(data_sub$pro))
buff_table[2,2] <- sum(abs(data_sub$ribo) > abs(data_sub$rna) &
                       abs(data_sub$pro) > abs(data_sub$rna))

buff_table
```


We observe that 39 out of 54 genes are buffered at the Ribo level
given that they are buffered at both Ribo and Protein level.


Here's why...
=============

Kushal was the mastermind behind the following solution. I think this is the right way of looking at this problem... Let us know if you think otherwise. 

*Question: Why is that table[1,1] and table[2,2] entries are always greater than table[1,2] and table[2,1] entries, even after our permutation of gene labels??*

We can ask this question this way: Given three objects A, B, and C, let's count the number of possible combinations of these three objects under the following scenarios:

* Scenario 1: A > B, A > C --> Makes 2 combinations
* Scenario 2: A > B, A < C --> equivalent to C > A > B --> 1 combination
* Scenario 3: A < B, A > C --> equivalent to C < A < B --> 1 combination
* Scenario 4: A < B, A < C --> Makes 2 combinations

In our table, [1,1] corresponds to Scenario 1 above and [1,2] corresponds to Scenario 2 above. The expectation of cell frequencies of the table is thus: 

```{r}
exp_table <- matrix(c(2,1,1,2), 2, 2, byrow = TRUE,
                    dimnames = list(RNA_Ribo = c("buffered", "non-buffered"),
                                    RNA_Pro = c("buffered", "non-buffered") ) )
exp_table
```







