---
output:
  knitrBootstrap::bootstrap_document:
    title: "Sanity check on TE divergence results"
    theme: default
    theme.chooser: TRUE
    highlight.chooser: TRUE
---


Checking TE divergence results 
=================================

```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


[view source](divergence-check-rna-ref.Rmd)


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
              bootstrap.panel = TRUE, 
              bootstrap.show.code = TRUE)
```


As a sanity check of our new definition of attenuation and enhancement of 
translation divergence relative to transcript divergence, we applied the same 
steps to compute the number of genes that are attenuated or enhanced at 
the transcript level compared to the translation level. We expect that the 
ratio of the number of genes that fall in to each category in transcript
attenuation or enhancement would the inverse of the ratio of the number of genes that fall into each category in translation attenuation or attenuation.


Definitions
============

The set of genes that are significant in inter-species difference of 
translation efficiency can be divided into two categories, and within
each of the categories, the genes can further be divided into two 
subcategories as follows

1. No sig. inter-species difference in Ribo

    1.1. No sig. inter-species difference in RNA
    
    1.2. Sig. inter-species difference in RNA
    
2. Sig. inter-species difference in Ribo

    2.1. ( abs(inter-species Ribo FC) > abs(inter-species RNA FC) ) OR
         ( inter-species RNA FC and inter-species Ribo FC are of 
           the opposite direction) - **Attenuation at the RNA level**
           
    2.2. ( abs(inter-species Ribo FC) < abs(inter-species RNA FC) ) AND
         ( inter-species RNA FC and inter-species Ribo FC are of the
           same direction) - **Enhancement at the RNA level**



```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```



 
Load rdas of LRT results and compute fold change estimates using
RPKM values before normalization. Note that RNA and Ribo data were 
quantile-normalized separately. 


```{r, bootstrap.show.code = TRUE}
## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))

## load LRT results of Ribo vs. RNA divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of DE analysis
load(file.path(rdadir,"DE.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load(file.path(rdadir,"eSetRRP.rda"))
eSet.temp = eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]
fc.mat=lapply(seq_along(c("human","chimp")), function(i) {
  ii = eSet.temp$species==c("human","chimp")[i]
  eSet.tt = eSet.temp[,ii]
  emat = lapply(seq_along(c("ribo","rna","protein")),function(j) {
    jj = eSet.tt$seqData==c("ribo","rna","protein")[j]
    rowMeans(exprs(eSet.tt[,jj]),na.rm=TRUE)
  })
  emat=do.call(cbind,emat)
  colnames(emat)=c("ribo","rna","protein")
  return(data.frame(emat))
})
names(fc.mat)=c("human","chimp")
dmat_unnormed = data.frame(ribo=fc.mat$human$ribo-fc.mat$chimp$ribo,
                           rna=fc.mat$human$rna-fc.mat$chimp$rna,
                           pro=fc.mat$human$protein-fc.mat$chimp$protein)

xy.rnapro = data.frame(rna=dmat_unnormed$rna,
                       pro=dmat_unnormed$pro)
xy.riborna = data.frame(rna=dmat_unnormed$rna,
                        ribo=dmat_unnormed$ribo)
xy.ribopro = data.frame(ribo=dmat_unnormed$ribo,
                        pro=dmat_unnormed$pro)
```


Define subsets of genes at FDR < .01.

```{r}
## Genes with sig. human/chimp difference in levels of translation efficiency 
ii_sig_TE <- res.riborna$int.qval < .01

## Genes with sig. human/chimp difference in RNA levels 
ii_sig_RNA_DE <- rnaRes.Q$qval < .01

## Genes with sig. human/chimp difference in Ribo levels
ii_sig_Ribo_DE <- riboRes.Q$qval < .01


## Category 1 - Enhanced divergence, criteria are as follows
## a) Sig. translation efficiency, AND
## b) Sig. DE at Ribo level, AND
## c) ( abs(RNA FC) > abs(Ribo FC) ) AND ( RNA FC * Ribo FC > 0)
iicheck_rnaGTribo <- 
    (ii_sig_TE) & ii_sig_Ribo_DE & 
    ( (abs(xy.riborna$rna) > abs(xy.riborna$ribo)) & 
      (xy.riborna$rna * xy.riborna$ribo > 0 ) )


## Category 2 - Attenuated divergence, criteria are as follows
## a) Sig. translation efficiency, AND
## b) Sig. DE at Ribo level, AND 
## c) ( abs(Ribo FC) > abs(RNA FC) ) AND ( RNA FC * Ribo FC < 0)
iicheck_riboGTrna <- 
    (ii_sig_TE) & ii_sig_Ribo_DE & 
    ( (abs(xy.riborna$ribo) > abs(xy.riborna$rna) ) | 
      (xy.riborna$rna * xy.riborna$ribo < 0) ) 


## Category 3, Pseudo enhacement, criteria are as follows
## a) Sig. translation efficiency, AND
## b) No sig. DE at Ribo level, AND 
## c) Sig. DE at RNA level
iicheck_null_rnaGTribo <- 
    ii_sig_TE & !ii_sig_Ribo_DE & ii_sig_RNA_DE


## Category 4, Null divergence, criteria are as follows
## a) Sig. translation efficiency, AND
## b) No sig. DE at Ribo level, AND 
## c) No sig. DE at RNA level
iicheck_null_rnaNULLribo <- 
    ii_sig_TE & !ii_sig_Ribo_DE & !ii_sig_RNA_DE


RNA_refmat <- rbind.data.frame(attenuated = sum(iicheck_riboGTrna),
                               enhanced = sum(iicheck_rnaGTribo),
                               null_enhanced = sum(iicheck_null_rnaGTribo),
                               null = sum(iicheck_null_rnaNULLribo) )
colnames(RNA_refmat) <- "# genes"
RNA_refmat
```






Session information
===========

```{r}
sessionInfo()
```

