---
output:
  knitrBootstrap::bootstrap_document:
    title: "Sanity check, relative to protein divergence"
    theme: default
    theme.chooser: TRUE
    highlight.chooser: TRUE
---


Sanity check, relative to protein divergence
============================================


```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


[view source](divergence-check-pro-ref-20150812.Rmd)


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
               bootstrap.panel = TRUE, 
               bootstrap.show.code = TRUE)
```



```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```

Goal
====

Count number of genes attenuated or enhanced at RNA level relative to protein leve.


Definitions
============

The set of genes that are significant in inter-species difference of 
RNA vs. Protein divergence can be divided into two categories, and within
each of the categories, the genes can further be divided into two 
subcategories as follows

1. ( abs(inter-species Protein FC) > abs(inter-species RNA FC) ) OR
   ( inter-species Protein FC and inter-species Ribo FC are of 
     the opposite direction) - **Attenuation at the RNA level**
     
2. ( abs(inter-species Protein FC) < abs(inter-species RNA FC) ) AND
   ( inter-species RNA FC and inter-species Protein FC are of the
     same direction) - **Enhancement at the RNA level**



```{r}
## Load rdas of LRT results and compute fold change estimates using
## RPKM values before normalization. 

## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load( file.path(rdadir,"eSetRRP.rda") )

eSet.temp <- eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]

fc.mat <- lapply( seq_along( c("human","chimp") ), function(i) {
  ii <- eSet.temp$species == c("human","chimp")[i]
  eSet.tt <- eSet.temp[, ii]
  emat <- lapply( seq_along( c("ribo","rna","protein") ), function(j) {
    jj <- eSet.tt$seqData == c("ribo","rna","protein")[j]
    rowMeans( exprs(eSet.tt[, jj] ), na.rm=TRUE )
  })
  emat <- do.call(cbind, emat)
  colnames(emat) <- c("ribo","rna","protein")
  return( data.frame(emat) )
})
names(fc.mat) <- c("human","chimp")

dmat_unnormed <- data.frame( ribo = fc.mat$human$ribo-fc.mat$chimp$ribo,
                             rna = fc.mat$human$rna-fc.mat$chimp$rna,
                             pro = fc.mat$human$protein-fc.mat$chimp$protein )

xy.rnapro <- data.frame( rna = dmat_unnormed$rna,
                         pro = dmat_unnormed$pro )
xy.riborna <- data.frame( rna = dmat_unnormed$rna,
                          ribo = dmat_unnormed$ribo )
xy.ribopro <- data.frame( ribo = dmat_unnormed$ribo,
                          pro = dmat_unnormed$pro )
```


Define subsets of genes at FDR < .01.

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and protein
ii_sig_rnapro <- res.rnapro$int.qval < .01

## Category 1 - Enhanced divergence, criteria are as follows
## a) Sig. RNA vs. Protein difference, AND
## b) ( abs(Protein FC) < abs(RNA FC) ) AND ( RNA FC * Protein FC > 0)
iicheck_rnaGTpro <- (ii_sig_rnapro) &  
                    ( (abs(xy.rnapro$rna) > abs(xy.rnapro$pro)) & 
                      (xy.rnapro$rna * xy.rnapro$pro > 0 ) )


## Category 2 - Attenuated divergence, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) ( abs(Protein FC) > abs(RNA FC) ) AND ( RNA FC * Protein FC < 0)
iicheck_proGTrna <- (ii_sig_rnapro) &  
                    ( (abs(xy.rnapro$pro) > abs(xy.rnapro$rna) ) | 
                      (xy.rnapro$rna * xy.rnapro$pro < 0) ) 



RNA_refmat <- rbind.data.frame( attenuated = sum(iicheck_proGTrna),
                                enhanced = sum(iicheck_rnaGTpro) )
colnames(RNA_refmat) <- "# genes"
kable(RNA_refmat)
```






Session information
===========

```{r}
sessionInfo()
```

