---
title: "include additional mm4 human data to correct batch effect"
author: "Sidney Wang"
date: "October 17, 2015"
output: html_document
---

Since most of rhesus and chimp are in mm4, while the human data include in this study are not, batch order is to some degree confounded with species. To correct for bach effect, we decided to first include the additional 4 human data from mm4 to regress out batch effect and then proceed with analysis on the original set of 16 samples. The analysis in this document will assess the validity of this approach.

I went back to process additional data. The file "process_additional_human_data_from_mm4_101515" documented detailed description on the process. I then save the data table as r object in the rdr folder

load data
```{r}
load("../rdas/HRC.plus.mm4H.ribo.rda")
load("../rdas/mm4.ribo.rda")
load("../rdas/HCR.ortho.geneLength.rda")
```

plot count range
```{r}
boxplot(log2(HRC.plus.mm4H.ribo[,-1]+1), outline =F)
boxplot(log2(mm4.ribo[,-1]+1), outline =F)
```

compute RPKM
=============

Since I found some slight variation in ortho genelength computed for each species, I will compute RPKM in the following 3 ways and compare resluts.
1. use human gene length
2. use the max length amongst the three species
3. use species specific gene length

use human gene length
```{r}
HRC.plus.mm4H.rpkm.hLength <- prop.table(as.matrix(HRC.plus.mm4H.ribo[,-1]),2)*10^9/HCR.geneLength$Human

mm4.rpkm.hLength <- prop.table(as.matrix(mm4.ribo[,-1]),2)*10^9/HCR.geneLength$Human
```

use max gene length
```{r}
HRC.plus.mm4H.rpkm.maxLength <- prop.table(as.matrix(HRC.plus.mm4H.ribo[,-1]),2)*10^9/HCR.geneLength$HCR.max

mm4.rpkm.maxLength <- prop.table(as.matrix(mm4.ribo[,-1]),2)*10^9/HCR.geneLength$HCR.max
```

use species specific length

```{r}
HRC.plus.mm4H.rpkm.speLength <- prop.table(as.matrix(HRC.plus.mm4H.ribo[,-1]),2)*10^9
HRC.plus.mm4H.rpkm.speLength <- cbind(HRC.plus.mm4H.rpkm.speLength[,1:5]/HCR.geneLength$Chimp,HRC.plus.mm4H.rpkm.speLength[,6:11]/HCR.geneLength$Human,HRC.plus.mm4H.rpkm.speLength[,12:16]/HCR.geneLength$Rhesus,HRC.plus.mm4H.rpkm.speLength[,17:20]/HCR.geneLength$Human)


mm4.rpkm.speLength <- prop.table(as.matrix(mm4.ribo[,-1]),2)*10^9
mm4.rpkm.speLength<-cbind(mm4.rpkm.speLength[,1:4]/HCR.geneLength$Chimp, mm4.rpkm.speLength[,5:8]/HCR.geneLength$Rhesus, mm4.rpkm.speLength[,9:12]/HCR.geneLength$Human)

```


filter data by requiring at least half of mm4 data have rpkm greater than 0

```{r}
rpkm.filter<-c() 
for (i in 1:length(mm4.rpkm.speLength[,1])){
rpkm.filter[i]<-sum(mm4.rpkm.speLength[i,] == 0) < 7
}

HRC.plus.mm4H.rpkm.speLength<-HRC.plus.mm4H.rpkm.speLength[rpkm.filter,]
mm4.rpkm.speLength<-mm4.rpkm.speLength[rpkm.filter,]
```

quantile normalize and log2 transform data
```{r}
library(limma)
HRC.plus.mm4H.rpkm.speLength.Q<-normalizeQuantiles(HRC.plus.mm4H.rpkm.speLength,ties = T)
mm4.rpkm.speLength.Q<-normalizeQuantiles(mm4.rpkm.speLength,ties = T)
log2.HRC.plus.mm4H.rpkm.speLength.Q<-log2(HRC.plus.mm4H.rpkm.speLength.Q)
is.na(log2.HRC.plus.mm4H.rpkm.speLength.Q)<-is.infinite(log2.HRC.plus.mm4H.rpkm.speLength.Q)
log2.mm4.rpkm.speLength.Q<-log2(mm4.rpkm.speLength.Q)
is.na(log2.mm4.rpkm.speLength.Q)<-is.infinite(log2.mm4.rpkm.speLength.Q)
```

DE test, LRT 

```{r}
species.label<-as.factor(substring(colnames(log2.mm4.rpkm.speLength.Q),1,1))
CRpvalue<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,1:8]~species.label[1:8])
reducedModel<-lm(log2.mm4.rpkm.speLength.Q[i,1:8]~1)
CRpvalue[i]<-anova(fullModel,reducedModel)$"Pr(>F)"[2]
}

HRpvalue<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,5:12]~species.label[5:12])
reducedModel<-lm(log2.mm4.rpkm.speLength.Q[i,5:12]~1)
HRpvalue[i]<-anova(fullModel,reducedModel)$"Pr(>F)"[2]
}

HCpvalue<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,-5:-8]~species.label[-5:-8])
reducedModel<-lm(log2.mm4.rpkm.speLength.Q[i,-5:-8]~1)
HCpvalue[i]<-anova(fullModel,reducedModel)$"Pr(>F)"[2]
}


```

qvalue 
```{r}
library(qvalue)
qvalue(HCpvalue)$pi0
qvalue(CRpvalue)$pi0
qvalue(HRpvalue)$pi0
length(which(qvalue(HCpvalue)$qvalues < 0.01))
length(which(qvalue(HRpvalue)$qvalues < 0.01))
length(which(qvalue(CRpvalue)$qvalues < 0.01))

vennDiagram(cbind(qvalue(HCpvalue)$qvalues < 0.01,qvalue(HRpvalue)$qvalues < 0.01,qvalue(CRpvalue)$qvalues < 0.01),names = c("HC","HR","CR"))
```


DE test, beta

```{r}
species.label<-as.factor(substring(colnames(log2.mm4.rpkm.speLength.Q),1,1))
CRpvalue<-c()
CR.effect <- c()
CR.se<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,1:8]~species.label[1:8])
CRpvalue[i]<-summary(fullModel)$coefficients[2,4]
CR.effect[i] <- summary(fullModel)$coefficients[2,1]
CR.se[i] <- summary(fullModel)$coefficients[2,2]

}

HRpvalue<-c()
HR.effect <- c()
HR.se <- c()

for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,5:12]~species.label[5:12])
HRpvalue[i]<-summary(fullModel)$coefficients[2,4]
HR.effect[i] <- summary(fullModel)$coefficients[2,1]
HR.se[i] <- summary(fullModel)$coefficients[2,2]

}

HCpvalue<-c()
HC.effect <- c()
HC.se <- c()

for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,-5:-8]~species.label[-5:-8])
HCpvalue[i]<-summary(fullModel)$coefficients[2,4]
HC.effect[i] <- summary(fullModel)$coefficients[2,1]
HC.se[i] <- summary(fullModel)$coefficients[2,2]

}


```


effect size and variance

```{r}

boxplot(log2(abs(HC.effect)),log2(abs(HR.effect)),log2(abs(CR.effect)), xlab="",main="beta",ylab="log2 absolute slope", notch=T)
axis(1,at = c(1,2,3),labels = c("HC","HR","CR"))

boxplot(log2(HC.se),log2(HR.se),log2(CR.se), xlab="",main="se",ylab="log2 SE", notch=T)
axis(1,at = c(1,2,3),labels = c("HC","HR","CR"))

```




pca
```{r}
library(ggplot2)
library(reshape2)
library(dplyr)
library(gdata)
pc<-prcomp(t(na.omit(log2.mm4.rpkm.speLength.Q)),center = T,scale = T)
barplot((pc$sdev)^2/sum((pc$sdev)^2), names.arg =  colnames(pc$x))
spec<-substring(rownames(pc$x),1,1)
temp.data<-as.data.frame(cbind(melt(pc$x),pc$x[,1],spec))
names(temp.data)<-c("ID","PC","value","pc1","species")

temp.data %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = species))+facet_wrap(~PC)


read.xls("/Users/siddisis/Dropbox/work/Ribosome_profiling/Primate_comparison/primate_ribo_gender_mmbatch.xlsx")->cell.source
cell.source<-cell.source[,1:4]


batch<-c(as.character(cell.source$libmix),rep("mm4",4))

pc<-prcomp(t(na.omit(log2.HRC.plus.mm4H.rpkm.speLength.Q)),center = T,scale = T)
barplot((pc$sdev)^2/sum((pc$sdev)^2), names.arg =  colnames(pc$x))
spec<-substring(rownames(pc$x),1,1)
temp.data<-as.data.frame(cbind(melt(pc$x),pc$x[,1],spec,batch))
names(temp.data)<-c("ID","PC","value","pc1","species","batch")

temp.data %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = species))+facet_wrap(~PC)
temp.data %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = batch))+facet_wrap(~PC)
temp.data %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = species, shape = batch))+facet_wrap(~PC)


```

within speceis SD
```{r}
SD.ribo.H<-apply(log2.mm4.rpkm.speLength.Q[,9:12], 1,sd,na.rm = T)
mean.ribo.H<-apply(log2.mm4.rpkm.speLength.Q[,9:12], 1,mean,na.rm = T)
SD.ribo.R<-apply(log2.mm4.rpkm.speLength.Q[,5:8], 1,sd,na.rm = T)
mean.ribo.R<-apply(log2.mm4.rpkm.speLength.Q[,5:8], 1,mean,na.rm = T)
SD.ribo.C<-apply(log2.mm4.rpkm.speLength.Q[,1:4], 1,sd,na.rm = T)
mean.ribo.C<-apply(log2.mm4.rpkm.speLength.Q[,1:4], 1,mean,na.rm = T)

CV.ribo.H <- SD.ribo.H - mean.ribo.H
CV.ribo.R <- SD.ribo.R - mean.ribo.R
CV.ribo.C <- SD.ribo.C - mean.ribo.C

boxplot(SD.ribo.H,SD.ribo.R,SD.ribo.C,xlab="", notch=T)
axis(1,c(1,2,3),c("human","rhesus","chimp"))

boxplot(CV.ribo.H,CV.ribo.R,CV.ribo.C,xlab="", notch=T)
axis(1,c(1,2,3),c("human","rhesus","chimp"))

```

weighted least squared

```{r}
Var.ribo.R <- SD.ribo.R^2
Var.ribo.C <- SD.ribo.C^2
Var.ribo.H <- SD.ribo.H^2
species.label<-as.factor(substring(colnames(log2.mm4.rpkm.speLength.Q),1,1))
CRpvalue.weighted.var<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,1:8]~species.label[1:8],weights = c(rep(1/Var.ribo.C[i], 4),rep(1/Var.ribo.R[i], 4)))
CRpvalue.weighted.var[i]<-summary(fullModel)$coefficients[2,4]
}

HRpvalue.weighted.var<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,5:12]~species.label[5:12],weights = c(rep(1/Var.ribo.R[i], 4),rep(1/Var.ribo.H[i], 4)))
HRpvalue.weighted.var[i]<-summary(fullModel)$coefficients[2,4]
}

HCpvalue.weighted.var<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,-5:-8]~species.label[-5:-8],weights = c(rep(1/Var.ribo.C[i], 4),rep(1/Var.ribo.H[i], 4)))

HCpvalue.weighted.var[i]<-summary(fullModel)$coefficients[2,4]
}



qvalue(HCpvalue.weighted.var)$pi0
qvalue(CRpvalue.weighted.var)$pi0
qvalue(HRpvalue.weighted.var)$pi0
length(which(qvalue(HCpvalue.weighted.var)$qvalues < 0.01))
length(which(qvalue(HRpvalue.weighted.var)$qvalues < 0.01))
length(which(qvalue(CRpvalue.weighted.var)$qvalues < 0.01))



CRpvalue.weighted.sd<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,1:8]~species.label[1:8],weights = c(rep(1/SD.ribo.C[i], 4),rep(1/SD.ribo.R[i], 4)))
CRpvalue.weighted.sd[i]<-summary(fullModel)$coefficients[2,4]
}

HRpvalue.weighted.sd<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,5:12]~species.label[5:12],weights = c(rep(1/SD.ribo.R[i], 4),rep(1/SD.ribo.H[i], 4)))
HRpvalue.weighted.sd[i]<-summary(fullModel)$coefficients[2,4]
}

HCpvalue.weighted.sd<-c()
for (i in 1:length(log2.mm4.rpkm.speLength.Q[,1])){
fullModel<-lm(log2.mm4.rpkm.speLength.Q[i,-5:-8]~species.label[-5:-8],weights = c(rep(1/SD.ribo.C[i], 4),rep(1/SD.ribo.H[i], 4)))

HCpvalue.weighted.sd[i]<-summary(fullModel)$coefficients[2,4]
}

qvalue(HCpvalue.weighted.sd)$pi0
qvalue(CRpvalue.weighted.sd)$pi0
qvalue(HRpvalue.weighted.sd)$pi0
length(which(qvalue(HCpvalue.weighted.sd)$qvalues < 0.01))
length(which(qvalue(HRpvalue.weighted.sd)$qvalues < 0.01))
length(which(qvalue(CRpvalue.weighted.sd)$qvalues < 0.01))

```


regress out batch (fix effect)

```{r}


res.list<-list()

for (i in 1:length(log2.HRC.plus.mm4H.rpkm.speLength.Q[,1])){
batch.fit<-lm(log2.HRC.plus.mm4H.rpkm.speLength.Q[i,]~as.factor(batch))
res.list[[i]]<-as.data.frame(t(batch.fit$residuals))
}
batch.fit.res<-rbind_all(res.list)
pc<-prcomp(t(na.omit(batch.fit.res)),center = T,scale = T)
#pc<-prcomp(batch.fit$residuals[1:13,],center = T,scale = T)

barplot((pc$sdev)^2/sum((pc$sdev)^2), names.arg =  colnames(pc$x))
spec<-substring(rownames(pc$x),1,1)
temp.data<-as.data.frame(cbind(melt(pc$x),pc$x[,1],spec,batch))
names(temp.data)<-c("ID","PC","value","pc1","species","batch")

temp.data %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = species, shape=batch))+facet_wrap(~PC)


```

protect main effect and regress out batch (fix effect) 

```{r}

spec<-substring(colnames(log2.HRC.plus.mm4H.rpkm.speLength.Q),1,1)
spec[17:20]<-c("H","H","H","H")
spec.design <- model.matrix(~ 0+spec)

res.list<-list()
intercept<-c()
beta1<-c()
beta2<-c()
for (i in 1:length(log2.HRC.plus.mm4H.rpkm.speLength.Q[,1])){
batch.fit<-lm(log2.HRC.plus.mm4H.rpkm.speLength.Q[i,]~as.factor(spec)+as.factor(batch))
res.list[[i]]<-as.data.frame(t(batch.fit$residuals)) 
intercept[i]<-batch.fit$coefficients[1]    
beta1[i]<-batch.fit$coefficients[2]
beta2[i]<-batch.fit$coefficients[3]

                          }

batch.fit.res<-rbind_all(res.list)
batch.removed<-batch.fit.res+intercept+beta1 %*% t(spec.design[,2])+beta2 %*% t(spec.design[,3])

pc<-prcomp(t(na.omit(batch.removed)),center = T,scale = T)
#pc<-prcomp(batch.fit$residuals[1:13,],center = T,scale = T)

barplot((pc$sdev)^2/sum((pc$sdev)^2), names.arg =  colnames(pc$x))
spec<-substring(rownames(pc$x),1,1)
temp.data<-as.data.frame(cbind(melt(pc$x),pc$x[,1],spec,batch))
names(temp.data)<-c("ID","PC","value","pc1","species","batch")

temp.data %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = species, shape=batch))+facet_wrap(~PC)


```



limma remove batch effect

```{r}

spec<-substring(colnames(log2.HRC.plus.mm4H.rpkm.speLength.Q),1,1)
spec[17:20]<-c("H","H","H","H")
spec.design <- model.matrix(~ 0+spec)
#batch.removed.limma.full<-removeBatchEffect(log2.HRC.plus.mm4H.rpkm.speLength.Q,batch = as.factor(batch))
batch.removed.limma.full<-removeBatchEffect(log2.HRC.plus.mm4H.rpkm.speLength.Q,batch = as.factor(batch),design = spec.design)


pc<-prcomp(t(na.omit(batch.removed.limma.full)),center = T,scale = T)
#pc<-prcomp(batch.fit$residuals[1:13,],center = T,scale = T)

barplot((pc$sdev)^2/sum((pc$sdev)^2), names.arg =  colnames(pc$x))
spec<-substring(rownames(pc$x),1,1)
temp.data<-as.data.frame(cbind(melt(pc$x),pc$x[,1],spec,batch))
names(temp.data)<-c("ID","PC","value","pc1","species","batch")

temp.data %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = species, shape=batch))+facet_wrap(~PC)


```



