---
output:
  knitrBootstrap::bootstrap_document:
    title: "Translation divergence and RNA-protein divergence"
    theme: default
---


Translation divergence and RNA-protein divergence
=================================================

```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
               bootstrap.panel = FALSE, bootstrap.show.code = FALSE)
```




Goal
=====

Of all the 3,188 genes included in the analysis, we investigate divergence at the protein level relative to the RNA level in genes that 1) are considered to be **attenuated** at the protein level relative to the RNA level - (sig. different between RNA and protein divergence) *AND* ( (the absolute value of protein divergence is smaller than the absolute value of RNA divergence) *OR* (protein divergence is in the opposite direction of RNA divergence) ); 2) are considered to be **enhanced** at the protein level relative to the RNA level - (sig. different between RNA and protein divergence) *AND* ( (the absolute value of protein divergence is greater than the absolute value of RNA divergence) *AND* (protein divergence is in the opposite direction of RNA divergence) ); 3) are neither enhanced nor attenuated at the protein level relative to the RNA level. 

We apply the above definition of enhanced and attenuated divergence to also the discussion of divergence at the translation level relative to the RNA level. 



Definitions
============

The set of genes that are significant in inter-species difference of 
RNA vs. Ribo (translation efficiency) can be divided into two categories, and within each of the categories, the genes can further be divided into two subcategories as follows

1. No sig. inter-species difference in RNA

    1.1. No sig. inter-species difference in Ribo
    
    1.2. Sig. inter-species difference in Ribo
    
2. Sig. inter-species difference in RNA

    2.1. ( abs(inter-species RNA FC) > abs(inter-species Ribo FC) ) OR
         ( inter-species RNA FC and inter-species Ribo FC are of 
           the opposite direction) - **Attenuation at the Ribo level**
           
    2.2. ( abs(inter-species RNA FC) < abs(inter-species Ribo FC) ) AND
         ( inter-species RNA FC and inter-species Ribo FC are of the
           same direction) - **Enhancement at the Ribo level**



```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```

Load rdas of LRT results and compute fold change estimates using
RPKM values before quantile normalization. 


```{r, bootstrap.show.code = TRUE}
## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))

## load LRT results of Ribo vs. RNA divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of DE analysis
load(file.path(rdadir,"DE.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load(file.path(rdadir,"eSetRRP.rda"))
eSet.temp = eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]
fc.mat=lapply(seq_along(c("human","chimp")), function(i) {
  ii = eSet.temp$species==c("human","chimp")[i]
  eSet.tt = eSet.temp[,ii]
  emat = lapply(seq_along(c("ribo","rna","protein")),function(j) {
    jj = eSet.tt$seqData==c("ribo","rna","protein")[j]
    rowMeans(exprs(eSet.tt[,jj]),na.rm=TRUE)
  })
  emat=do.call(cbind,emat)
  colnames(emat)=c("ribo","rna","protein")
  return(data.frame(emat))
})
names(fc.mat)=c("human","chimp")
dmat_unnormed = data.frame(ribo=fc.mat$human$ribo-fc.mat$chimp$ribo,
                           rna=fc.mat$human$rna-fc.mat$chimp$rna,
                           pro=fc.mat$human$protein-fc.mat$chimp$protein)

xy.rnapro = data.frame(rna=dmat_unnormed$rna,
                       pro=dmat_unnormed$pro)
xy.riborna = data.frame(rna=dmat_unnormed$rna,
                        ribo=dmat_unnormed$ribo)
xy.ribopro = data.frame(ribo=dmat_unnormed$ribo,
                        pro=dmat_unnormed$pro)
```


Define subsets of genes at FDR < .01.

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and Ribo
ii_sig_rnaribo <- res.riborna$int.qval < .01

## Genes with sig. human/chimp difference in divergence between RNA and Protein
ii_sig_rnapro <- res.rnapro$int.qval < .01

## Genes with sig. human/chimp difference in RNA levels
ii_sig_RNA_DE <- rnaRes.Q$qval < .01

## Genes with sig. human/chimp difference in Ribo levels
ii_sig_Ribo_DE <- riboRes.Q$qval < .01

## Genes with sig. human/chimp difference in Ribo levels
ii_sig_Pro_DE <- massRes.Q$qval < .01
```



## Define RNA vs. Protein divergence

```{r}
## Category 1 - Enhanced divergence, criteria are as follows
## a) Sig. RNA vs. Protein difference, AND
## b) Sig. DE at RNA level, AND
## c) ( abs(RNA FC) < abs(Protein FC) ) AND ( RNA FC * Protein FC > 0)
ii_proGTrna <- 
    (ii_sig_rnapro) & ii_sig_RNA_DE & 
    ( (abs(xy.rnapro$pro) > abs(xy.rnapro$rna)) & 
      (xy.rnapro$rna * xy.rnapro$pro > 0 ) )


## Category 2 - Attenuated divergence, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) Sig. DE at RNA level, AND 
## c) ( abs(RNA FC) > abs(Protein FC) ) AND ( RNA FC * Protein FC < 0)
ii_rnaGTpro <- 
    (ii_sig_rnapro) & ii_sig_RNA_DE & 
    ( (abs(xy.rnapro$rna) > abs(xy.rnapro$pro) ) | 
      (xy.rnapro$rna * xy.rnapro$pro < 0) ) 


## Category 3, Pseudo enhacement, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) No sig. DE at RNA level, AND 
## c) Sig. DE at Protein level
ii_null_rnaGTpro <- 
    ii_sig_rnapro & !ii_sig_RNA_DE & ii_sig_Pro_DE


## Category 4, Null divergence, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) No sig. DE at RNA level, AND 
## c) No sig. DE at Protein level
ii_null_rnaNULLprotein <- 
    ii_sig_rnapro & !ii_sig_RNA_DE & !ii_sig_Pro_DE


Protein_refmat <- rbind.data.frame(attenuated = sum(ii_rnaGTpro),
                                   enhanced = sum(ii_proGTrna),
                                   null_enhanced = sum(ii_null_rnaGTpro),
                                   null = sum(ii_null_rnaNULLprotein) )
colnames(Protein_refmat) <- "# genes"
Protein_refmat
```


## Define RNA vs. Ribo divergence

```{r}
## Category 1 - Enhanced divergence, criteria are as follows
## a) Sig. RNA vs. Ribo difference, AND
## b) Sig. DE at RNA level, AND
## c) ( abs(RNA FC) < abs(Ribo FC) ) AND ( RNA FC * Ribo FC > 0)
ii_riboGTrna <- 
    (ii_sig_rnaribo) & ii_sig_RNA_DE & 
    ( (abs(xy.riborna$ribo) > abs(xy.riborna$rna)) & 
      (xy.riborna$rna * xy.riborna$ribo > 0 ) )


## Category 2 - Attenuated divergence, criteria are as follows
## a) Sig. RNA vs. Ribo diff., AND
## b) Sig. DE at RNA level, AND 
## c) ( abs(RNA FC) > abs(Ribo FC) ) AND ( RNA FC * Ribo FC < 0)
ii_rnaGTribo <- 
    (ii_sig_rnaribo) & ii_sig_RNA_DE & 
    ( (abs(xy.riborna$rna) > abs(xy.riborna$ribo) ) | 
      (xy.riborna$rna * xy.riborna$ribo < 0) ) 


## Category 3, Pseudo enhacement, criteria are as follows
## a) Sig. RNA vs. Ribo diff., AND
## b) No sig. DE at RNA level, AND 
## c) Sig. DE at Ribo level
ii_null_rnaGTribo <- 
    ii_sig_rnaribo & !ii_sig_RNA_DE & ii_sig_Ribo_DE


## Category 4, Null divergence, criteria are as follows
## a) Sig. RNA vs. Ribo diff., AND
## b) No sig. DE at RNA level, AND 
## c) No sig. DE at Ribo level
ii_null_rnaNULLribo <- 
    ii_sig_rnaribo & !ii_sig_RNA_DE & !ii_sig_Ribo_DE


Ribo_refmat <- rbind.data.frame(attenuated = sum(ii_rnaGTribo),
                                   enhanced = sum(ii_riboGTrna),
                                   null_enhanced = sum(ii_null_rnaGTribo),
                                   null = sum(ii_null_rnaNULLribo) )
colnames(Ribo_refmat) <- "# genes"
Ribo_refmat
```



RNA ~ Protein
=============

Among the ``r sum( !(res.rnapro$int.qval < .01 & rnaRes.Q$qval < .01) )`` genes that are neither enhanced nor attenuated at the protein level compared to the RNA level (RNA ~ Protein), ``r sum( !(res.rnapro$int.qval < .01 & rnaRes.Q$qval < .01)  & (res.riborna$int.qval < .01) )`` genes are significantly different in divergence between RNA and Ribo. We computed the number of genes that are enhanced or attenuated at the translation level among these ``r sum( !(res.rnapro$int.qval < .01 & rnaRes.Q$qval < .01)  & (res.riborna$int.qval < .01) )`` genes.  

1. Attenuated translation: (sig. difference between RNA and Ribo) AND ( abs(RNA divergence) > abs(Ribo divergence) OR Ribo divergence is in the opposite direction of RNA divergence) AND (sig. RNA DE)

2. Reinforced translation: (sig. difference between RNA and Ribo) AND (abs(RNA divergence) < abs(Ribo divergence) AND Ribo divergence is in the same direction as RNA divergence) AND (sig. RNA DE)


```{r}
ii_null_rnapro <- !(res.rnapro$int.qval < .01 & rnaRes.Q$qval < .01)  

## RNA ~ Protein 
boxplot(abs(dmat_unnormed$rna[ii_null_rnapro]),
        abs(dmat_unnormed$ribo[ii_null_rnapro]),
        abs(dmat_unnormed$pro[ii_null_rnapro]), 
        axes = F, ylim = c(0, 10))
axis(1, at = c(1,2,3), labels = c("RNA", "Ribo", "Protein"), 
     col = "white")
axis(2)
points_cols <- c("Scarlet", "Sunglow", "Shamrock")
points(rep(1, sum(ii_null_rnapro)), 
       abs(dmat_unnormed$rna[ii_null_rnapro]), 
       pch = 19, cex = .5, 
       col = crayon[points_cols[1]] )
points(rep(2, sum(ii_null_rnapro)), 
       abs(dmat_unnormed$ribo[ii_null_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[2]] )
points(rep(3, sum(ii_null_rnapro)), 
       abs(dmat_unnormed$pro[ii_null_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[3]] )
text(2, 10, adj = 0,
     label== paste("RNA ~ Protein", sum(ii_null_rnapro), "genes" ) )
text(2, 9, adj = 0,
     label = paste("RNA vs. Ribo sig. diff.", 
       sum(ii_proNULLrna & (res.riborna$int.qval<.01)), "genes; \n",
       round(100*sum(ii_proNULLrna & (res.riborna$int.qval<.01))/
               sum(ii_proNULLrna), 2), "%") )
text(2, 8, adj = 0,
     label = paste("Attenuated translation", 
       sum(ii_attenuated_set), "genes; \n",
       round(100*sum(ii_attenuated_set)/sum(ii_proNULLrna), 2), "%") )
text(2, 7, adj = 0,
     label = paste("Enhanced translation", 
       sum(ii_enhanced_set), "genes; \n",
       round(100*sum(ii_enhanced_set)/sum(ii_proNULLrna), 2), "%") )
title(main = "RNA ~ Protein, no sig. diff. divergence",
      ylab = "absolute divergence (human-chimpanzee fold change)")

# # Compute pairwise p-values between effect sizes of Ribo, RNA, and protein
# # among genes with RNA ~ Protein
# rna_ribo_t <- t.test( abs(dmat_unnormed$rna[ii_proNULLrna]),
#        abs(dmat_unnormed$ribo[ii_proNULLrna]), paired = TRUE )
# rna_pro_t <- t.test( abs(dmat_unnormed$rna[ii_proNULLrna]),
#        abs(dmat_unnormed$pro[ii_proNULLrna]), paired = TRUE )
# ribo_pro_t <- t.test( abs(dmat_unnormed$ribo[ii_proNULLrna]),
#        abs(dmat_unnormed$pro[ii_proNULLrna]), paired = TRUE )
# 
# text(.5, 10, adj = 0,
#      label = paste("Paired t-test of effect sizes") )
# text(.5, 9, adj = 0,
#      label = paste("RNA vs. Ribo: p-value < 1e-14") )
# text(.5, 8, adj = 0,
#      label = paste("RNA vs. Protein: p-value < 1e-15") )
# text(.5, 7, adj = 0,
#      label = paste("Ribo vs. Protein: p-value < 1e-15") )
# 

```



### Signal in RNA ~ Protein

> Among the 408 genes with RNA and Ribo significant differences and no 
significant difference between RNA and protein, we do a follow-up analysis 
to look further to consider the contribution/relation of translation divergence 
to protein divergence, relative to RNA divergence. 

> Yoav: So, for example, if gene 1 (among the 408 genes with sig. diff. 
between RNA and Ribo) shows attenuated translation relative to RNA, protein 
divergence will be attenuated relative to RNA divergence. If gene 2 (among the 
408 genes with sig. diff. between RNA and Ribo) shows enhanced ribo divergence 
compared with RNA, I predict that the protein divergence will also be enhanced 
compared with RNA.


```{r}
ii_sig_RNA_DE <- rnaRes.Q$qval < .01
ii_enhanced_set_rna_ribo <- 
  ( (dmat_unnormed$rna * dmat_unnormed$ribo > 0) & 
  ( abs(dmat_unnormed$ribo) > abs(dmat_unnormed$rna) ) ) & 
  (res.riborna$int.qval < .01) & ii_sig_RNA_DE & ii_proNULLrna
ii_attenuated_set_rna_ribo <- 
  ( (dmat_unnormed$rna * dmat_unnormed$ribo < 0) | 
  ( abs(dmat_unnormed$ribo) < abs(dmat_unnormed$rna) ) ) & 
  (res.riborna$int.qval < .01) & ii_sig_RNA_DE & ii_proNULLrna



plot_mat <- matrix(0, 2, 2)
rownames(plot_mat) <- c("RNA-Ribo enhanced", "RNA-Ribo attenuated")
colnames(plot_mat) <- c("RNA-Protein enhanced", "RNA-Protein attenuated")

plot_mat[1,1] <- sum(ii_enhanced_set_rna_ribo*ii_enhanced_set)
plot_mat[1,2] <- sum(ii_enhanced_set_rna_ribo*ii_attenuated_set)
plot_mat[2,1] <- sum(ii_attenuated_set_rna_ribo*ii_enhanced_set)
plot_mat[2,2] <- sum(ii_attenuated_set_rna_ribo*ii_attenuated_set)

kable(plot_mat)
```


#### Attenuated translation

>  Among the 62 genes that were attenuated at the translation level relative to the transcript level (RNA-Ribo attenuation), we tested the null hypothesis that these genes are as likely to be attenuated as to be enhanced at the protein level relative to the transcript level (RNA-Protein). If this null hypothesis were rejected and the number of genes that were attenuated at the protein layer is larger than the number of genes that were enhanced at the protein layer, then we say that attenuation at the translation level is more likey to co-occur with attenuation at the protein level than with enhancement at the protein level.

> *Results*: YES (p-value < 1e-15)! Genes that were attenuated at the translation level are more likely to be attenuated at the protein level than to be enhanced at the protein level. 93% of the genes with attenuated translation relative to the transcript level are also attenuated at the protein level (relative to the transcript level).

```{r}
binom.test(plot_mat[2,2], rowSums(plot_mat)[2])
```


#### Enhanced translation

>  Among the 9 genes that were reinforced at translation, we tested the null hypothesis that these genes are as likely to be enhanced as to be attenuated at the protein layer. If this null hypothesis were rejected and the number of genes that were enhanced at the protein layer is more than those attenuated at the protein layer, then we say that genes that are enhanced at the translation level are more likely to be enhanced at the protein level than to be attenuated at the protein level.

> *Results*: NO (p-value = 1)! Genes that are enhanced at the translation level (relative to RNA) are no more likely to be enhanced at the protein level than to be attenuated at the protein level. 46% of the genes that are enhanced at the translation level are also enhanced at the protein level, relative to the transcript level of species divergence.

```{r}
binom.test(plot_mat[1,1], rowSums(plot_mat)[1])
```





RNA > Protein
==============

Among the genes that are attenuated at the protein level compared to the RNA level (RNA > protein), we are interested in the number of genes with attenuated translation as defined by the following: 1) sig. diff between Ribo and RNA, AND 2) when direction of species effect at Ribo is flipped in comparison to RNA regardless of the magnitude of the effect size *OR* when there is less divergence at the Ribo level and at the RNA level.

> Significant enrichment (p-value < 1e-15). Genes that are attenuated at the protein level are more likely to be attenuated at the translation level than to be enhanced at the translation level. The test is performend against the null hypothesis of prop. of attenuated genes under RNA ~ Protein (285/2877 = .0991).  


```{r}
# ii_subset1 <- dmat_unnormed$rna*dmat_unnormed$ribo < 0
# ii_subset2 <- abs(dmat_unnormed$ribo) < abs(dmat_unnormed$rna)
# ii_sub <- ii_subset1 | ii_subset2
# ii_attenuated_translation <- (res.riborna$int.qval<.01)*ii_sub

## RNA > Protein 
index_list <- list(ii_proNULLrna, ii_rnaGTpro, ii_proGTrna)
boxplot(abs(dmat_unnormed$rna[ii_rnaGTpro]),
        abs(dmat_unnormed$ribo[ii_rnaGTpro]),
        abs(dmat_unnormed$pro[ii_rnaGTpro]), 
        axes = F, ylim = c(0, 10))
axis(1, at = c(1,2,3), labels = c("RNA", "Ribo", "Protein"), 
     col = "white")
axis(2)
points_cols <- c("Scarlet", "Sunglow", "Shamrock")
points(rep(1, sum(ii_rnaGTpro)), abs(dmat_unnormed$rna[ii_rnaGTpro]), 
       pch = 19, cex = .5, 
       col = crayon[points_cols[1]] )
points(rep(2, sum(ii_rnaGTpro)), abs(dmat_unnormed$ribo[ii_rnaGTpro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[2]] )
points(rep(3, sum(ii_rnaGTpro)), abs(dmat_unnormed$pro[ii_rnaGTpro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[3]] )
text(2.2, 9, adj = 0,
     label = paste("RNA > Protein", sum(ii_rnaGTpro), 
                   "genes" ) )
text(2.2, 8, adj = 0,
     label = paste("RNA vs. Ribo sig. diff.", 
      sum(ii_rnaGTpro*(res.riborna$int.qval<.01)), 
      "genes; \n",
      round(100*sum(ii_rnaGTpro*(res.riborna$int.qval<.01))/
            sum(ii_rnaGTpro), 2),"%") )
text(2.2, 7, adj = 0,
     label = paste("Attenuated at translation", 
      sum(ii_rnaGTpro*ii_rnaGTribo), "genes; \n",
      round(100*sum(ii_rnaGTpro*ii_rnaGTribo)/sum(ii_rnaGTpro), 2),"%") )
title(main = paste("RNA > Protein",sum(ii_rnaGTpro),
                   "genes (Attenuated divergence)"),
      ylab = "absolute divergence (human-chimpanzee fold change)")

binom.test( sum(ii_rnaGTpro*ii_rnaGTribo),
           sum(ii_rnaGTpro), p = 61/66)
```





## RNA < Protein

> Among the genes that are enhanced at the protein level relative to the RNA level (RNA < Protein), we are interested in the number of genes that are attenuated at the translation level relative to the RNA level: 1) sig. diff between Ribo and RNA, AND 2) when direction of species effect at Ribo is consistent with RNA, or when there is greater divergence effect size (absolute value of divergence) at the Ribo level than at the RNA level.

> Significant enrichment (p-value < .0005). Genes that are enhanced at the protein level are more likely to be enhanced at the translation level than to be attenuated at the translation level. The test is performend against the null hypothesis of prop. of enhanced genes under RNA ~ Protein (90/2877 = .0313).  

```{r}
# ii_subset1 <- dmat_unnormed$rna * dmat_unnormed$ribo > 0
# ii_subset2 <- abs(dmat_unnormed$ribo) > abs(dmat_unnormed$rna)
# ii_sub <- ii_subset1 & ii_subset2
# ii_sam <- ii_proGTrna * (res.riborna$int.qval<.01) * ii_subset1
# ii_reinforced_translation <- (res.riborna$int.qval<.01)*ii_sub

## RNA < Protein 
index_list <- list(ii_proNULLrna, ii_rnaGTpro, ii_proGTrna)
boxplot(abs(dmat_unnormed$rna[ii_proGTrna]),
        abs(dmat_unnormed$ribo[ii_proGTrna]),
        abs(dmat_unnormed$pro[ii_proGTrna]), 
        axes = F, ylim = c(0, 10))
axis(1, at = c(1,2,3), labels = c("RNA", "Ribo", "Protein"), 
     col = "white")
axis(2)
points_cols <- c("Scarlet", "Sunglow", "Shamrock")
points(rep(1, sum(ii_proGTrna)), abs(dmat_unnormed$rna[ii_proGTrna]), 
       pch = 19, cex = .5, 
       col = crayon[points_cols[1]] )
points(rep(2, sum(ii_proGTrna)), abs(dmat_unnormed$ribo[ii_proGTrna]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[2]] )
points(rep(3, sum(ii_proGTrna)), abs(dmat_unnormed$pro[ii_proGTrna]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[3]] )
text(1.5, 9, adj = 0,
     label = paste("Protein > RNA", sum(ii_proGTrna), "genes" ) )
text(1.5, 8, adj = 0,
     label = paste("RNA vs. Ribo sig. diff.", 
       sum(ii_proGTrna*(res.riborna$int.qval<.01)), "genes; \n",
       round(100*sum(ii_proGTrna*(res.riborna$int.qval<.01))/
             sum(ii_proGTrna), 2), "%") ) 
# text(1.5, 7, adj = 0,
#      label = paste("RNA vs. Ribo sig. diff. & same direction", 
#       sum(ii_sam), "genes; \n",
#       round(100*sum(ii_sam)/sum(ii_proGTrna),
#             2),"%") )
text(1.5, 6, adj = 0,
     label = paste("Reinfored at translation", 
      sum(ii_proGTrna*ii_riboGTrna), "genes; \n",
      round(100*sum(ii_proGTrna*ii_riboGTrna)/sum(ii_proGTrna),
            2),"%") )
title(main = paste("RNA < Protein",sum(ii_proGTrna),
                  "genes (Enhanced protein divergence)"),
      ylab = "absolute divergence (human-chimpanzee fold change)")



binom.test(sum(ii_proGTrna*ii_riboGTrna), 
           sum(ii_proGTrna), p = 4/5)
```



Session information
===================

```{r}
sessionInfo()
```



