---
output:
  knitrBootstrap::bootstrap_document:
    title: "RNA-Ribo divergence and RNA-Protein divergence"
    theme: default
---


RNA-Ribo divergence and RNA-Protein divergence
==============================================


```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```

[view source](rna-ribo-pro-buffering.Rmd)


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
               bootstrap.panel = TRUE, bootstrap.show.code = TRUE)
```


```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```


Goal
====

Compute number of genes buffered at the translation level and buffered at the 
protein level relative to divergence at the transcript level.


Definitions
===========

Of all the 3,188 genes included in the analysis, we classifed genes into
**buffered** and **non-buffered** at the translation level or at the
protein level as follows

1. RNA-Ribo divergnece

    1.1. **Buffered** at the translation level relative to the RNA level: 
          sig. difference in translation efficiency AND
          abs(RNA divergence) > abs(Ribo divergence)
          
    1.2. **Non-buffered** at the translation level relative to the RNA level:
          sig. difference in translation efficiency AND
          abs(RNA divergence) < abs(Ribo divergence)
    
2. RNA-Protein divergnece

    2.1. **Buffered** at the protein level relative to the RNA level: 
          sig. difference in divergence between RNA and protein AND
          abs(RNA divergence) > abs(protein divergence)
          
    2.2. **Non-buffered** at the protein level relative to the RNA level:
          sig. difference in divergence between RNA and protein AND
          abs(RNA divergence) < abs(protein divergence)

    


```{r}
## Load rdas of LRT results and compute fold change estimates using
## RPKM values before normalization. 

## load LRT results of RNA vs. Ribo divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load( file.path(rdadir,"eSetRRP.rda") )

eSet.temp <- eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]

fc.mat <- lapply( seq_along( c("human","chimp") ), function(i) {
  ii <- eSet.temp$species == c("human","chimp")[i]
  eSet.tt <- eSet.temp[, ii]
  emat <- lapply( seq_along( c("ribo","rna","protein") ), function(j) {
    jj <- eSet.tt$seqData == c("ribo","rna","protein")[j]
    rowMeans( exprs(eSet.tt[, jj] ), na.rm=TRUE )
  })
  emat <- do.call(cbind, emat)
  colnames(emat) <- c("ribo","rna","protein")
  return( data.frame(emat) )
})
names(fc.mat) <- c("human","chimp")

dmat_unnormed <- data.frame( ribo = fc.mat$human$ribo-fc.mat$chimp$ribo,
                             rna = fc.mat$human$rna-fc.mat$chimp$rna,
                             pro = fc.mat$human$protein-fc.mat$chimp$protein )

xy.rnapro <- data.frame( rna = dmat_unnormed$rna,
                         pro = dmat_unnormed$pro )
xy.riborna <- data.frame( rna = dmat_unnormed$rna,
                          ribo = dmat_unnormed$ribo )
xy.ribopro <- data.frame( ribo = dmat_unnormed$ribo,
                          pro = dmat_unnormed$pro )
```



Gene classification
===================

## RNA vs. Ribo divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and Ribo
ii_sig_rnaribo <- res.riborna$int.qval < .01


## Buffered divergence
## a) Sig. RNA vs. Ribo difference, AND
## b) ( abs(RNA FC) > abs(Ribo FC) ) 
ii_rnaGTribo <- (ii_sig_rnaribo) & 
                ( abs(xy.riborna$rna) > abs(xy.riborna$ribo) ) 

## Non-buffered divergence
## a) Sig. RNA vs. Ribo diff., AND
## b) ( abs(RNA FC) < abs(Ribo FC) ) 
ii_riboGTrna <- (ii_sig_rnaribo) &  
                ( abs(xy.riborna$rna) < abs(xy.riborna$ribo) ) 

Ribo_refmat <- rbind.data.frame(buffered = sum(ii_rnaGTribo),
                                nonbuffered = sum(ii_riboGTrna) )
colnames(Ribo_refmat) <- "# genes"
kable(Ribo_refmat)
```


## RNA vs. Protein divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and protein
ii_sig_rnapro <- res.rnapro$int.qval < .01

## Buffered divergence
## a) Sig. RNA vs. Protein difference, AND
## b) ( abs(RNA FC) > abs(Protein FC) ) 
ii_rnaGTpro <- (ii_sig_rnapro) &  
               ( abs(xy.rnapro$rna) > abs(xy.rnapro$pro) ) 
                 


## Non-buffered divergence
## a) Sig. RNA vs. Protein diff., AND
## b) ( abs(RNA FC) < abs(Protein FC) ) 
ii_proGTrna <- (ii_sig_rnapro) &  
               ( abs(xy.rnapro$rna) < abs(xy.rnapro$pro) ) 


Protein_refmat <- rbind.data.frame( buffered = sum(ii_rnaGTpro),
                                    nonbuffered = sum(ii_proGTrna) )
colnames(Protein_refmat) <- "# genes"
kable(Protein_refmat)
```




Session information
===================

```{r}
sessionInfo()
```

