---
output:
  knitrBootstrap::bootstrap_document:
    title: "RNA-Protein divergence conditioned on RNA-Ribo divergence"
    theme: default
---


RNA-Protein divergence conditioned on RNA-Ribo divergence
=========================================================

```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```

[view source](TE-genes-ribo-pro-buffering.Rmd)


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
               bootstrap.panel = TRUE, 
               bootstrap.show.code = TRUE)
```



```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```


Goal
=====

Consider the set of genes with sig. difference in translation efficiency 
between human and chimpanzee. We then take the set of genes that are 
buffered at the translation level and count the number of genes that are
classifed as buffered at the protein level. We apply the same steps for
non-buffered genes at the translation level.

These analyses are for investigating whether divergence at the translation
level is likely to be carried over to the protein level.


Problem
=======

For the definition of buffering an non-buffering genes, we consider the 
following two approaches:

* **Interaction approach**: Genes are classifed as buffered at the 
    translation or at the protein level if there is a
    significant interaction between Ribo or protein and RNA inter-species
    differences and if the effect size of the inter-species difference
    at the RNA level is greater than the effect size of the inter-species
    differene at the Ribo or protein level.
    
* **Size approach**: Genes are classifed as buffered at the translation
    or at the protein level if the effect size of the inter-species
    difference at the RNA level is greater than the
    effect size of the inter-species difference at the Ribo or protein level.


Definitions
===========

Take the comparison between RNA and Ribo for example, we describe the 
definition of buffered and non-buffered genes as follows:

* **Interaction approach**: 

    1. Buffered at Ribo level: Sig. RNA-Ribo interaction AND 
        ( abs(RNA divergence) > abs(Ribo divergence) )
        
    2. Non-buffered at Ribo leve: Sig. RNA-Ribo interaction AND
        ( abs(RNA divergence) < abs(Ribo divergence) )


* **Size approach**: 

    1. Buffered at Ribo level: ( abs(RNA divergence) > abs(Ribo divergence) )
        
    2. Non-buffered at Ribo leve: ( abs(RNA divergence) < abs(Ribo divergence) )


```{r}
## Load rdas of LRT results and compute fold change estimates using
## RPKM values before normalization. 

## load LRT results of RNA vs. Ribo divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load( file.path(rdadir,"eSetRRP.rda") )

eSet.temp <- eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]

fc.mat <- lapply( seq_along( c("human","chimp") ), function(i) {
  ii <- eSet.temp$species == c("human","chimp")[i]
  eSet.tt <- eSet.temp[, ii]
  emat <- lapply( seq_along( c("ribo","rna","protein") ), function(j) {
    jj <- eSet.tt$seqData == c("ribo","rna","protein")[j]
    rowMeans( exprs(eSet.tt[, jj] ), na.rm=TRUE )
  })
  emat <- do.call(cbind, emat)
  colnames(emat) <- c("ribo","rna","protein")
  return( data.frame(emat) )
})
names(fc.mat) <- c("human","chimp")

dmat_unnormed <- data.frame( ribo = fc.mat$human$ribo-fc.mat$chimp$ribo,
                             rna = fc.mat$human$rna-fc.mat$chimp$rna,
                             pro = fc.mat$human$protein-fc.mat$chimp$protein )

xy.rnapro <- data.frame( rna = dmat_unnormed$rna,
                         pro = dmat_unnormed$pro )
xy.riborna <- data.frame( rna = dmat_unnormed$rna,
                          ribo = dmat_unnormed$ribo )
xy.ribopro <- data.frame( ribo = dmat_unnormed$ribo,
                          pro = dmat_unnormed$pro )
```


## RNA vs. Ribo divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and Ribo
ii_sig_rnaribo <- res.riborna$int.qval < .01

## Genes with RNA > Ribo in effect sizes
ii_rnaGTribo <- abs(xy.riborna$rna) > abs(xy.riborna$ribo)

## Genes with RNA < Ribo in effect sizes
ii_riboGTrna <- abs(xy.riborna$ribo) > abs(xy.riborna$rna)
```


## RNA vs. Protein divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and protein
ii_sig_rnapro <- res.rnapro$int.qval < .01

## Genes with RNA > Protein in effect sizes
ii_rnaGTpro <- abs(xy.rnapro$rna) > abs(xy.rnapro$pro)

## Genes with RNA < Protein in effect sizes
ii_proGTrna <- abs(xy.rnapro$pro) > abs(xy.rnapro$rna)
```




Buffering at Ribo and Protein
==============================


## Interaction approach

```{r}
buffering_matrix_interact <- 
    matrix(0, 4, 4,
     dimnames = list(rnaribo = c("Ribo buffered & RNA-Ribo interaction", 
                                 "Ribo non-buffered & RNA-Ribo interaction",
                                 "RNA-Ribo no interaction",
                                 "Column sum"),
                   rnapro = c("Protein buffered & RNA-Protein interaction",
                              "Protein non-buffered & RNA-protein interaction",
                              "RNA-Protein no interaction",
                              "Row sum") ) )

## Among the genes buffered at translation,
## count the number of genes buffered at protein,
## non-buffered at protein, no interaction between RNA and protein,
## and the total number
buffering_matrix_interact[1,1] <- 
    sum( (ii_rnaGTpro & ii_sig_rnapro)[ii_rnaGTribo & ii_sig_rnaribo])
buffering_matrix_interact[1,2] <- 
    sum( (ii_proGTrna & ii_sig_rnapro)[ii_rnaGTribo & ii_sig_rnaribo])
buffering_matrix_interact[1,3] <- 
    sum( (!ii_sig_rnapro)[ii_rnaGTribo & ii_sig_rnaribo] )
buffering_matrix_interact[1,4] <- 
    sum(ii_rnaGTribo & ii_sig_rnaribo)


## Among the genes non-buffered at translation,
## count the number of genes buffered at protein,
## non-buffered at protein, no interaction between RNA and protein,
## and the total number
buffering_matrix_interact[2,1] <- 
    sum( (ii_rnaGTpro & ii_sig_rnapro)[ii_riboGTrna & ii_sig_rnaribo])
buffering_matrix_interact[2,2] <- 
    sum( (ii_proGTrna & ii_sig_rnapro)[ii_riboGTrna & ii_sig_rnaribo])
buffering_matrix_interact[2,3] <- 
  sum( (!ii_sig_rnapro)[ii_riboGTrna & ii_sig_rnaribo] )
buffering_matrix_interact[2,4] <- 
  sum(ii_riboGTrna & ii_sig_rnaribo)


## Of all genes with sig. difference in translation efficiency,
## count the number of genes buffered at protein,
## non-buffered at protein, and no interaction between RNA and protein
buffering_matrix_interact[4,1] <- 
    sum( (ii_rnaGTpro & ii_sig_rnapro)[ 
        (ii_rnaGTribo | ii_riboGTrna) & ii_sig_rnaribo] )
buffering_matrix_interact[4,2] <- 
    sum( (ii_proGTrna & ii_sig_rnapro)[ 
        (ii_rnaGTribo | ii_riboGTrna) & ii_sig_rnaribo ] )
buffering_matrix_interact[4,3] <- 
  sum((!ii_sig_rnapro)[ ii_sig_rnaribo] )

## No. of genes with sig. translation efficiency
buffering_matrix_interact[4,4] <- sum(ii_sig_rnaribo)

kable(buffering_matrix_interact)
```



For genes buffered at the Ribo level compared to the RNA level, 
plot out per gene effect sizes of RNA and protein for genes
that are buffered at protein, non-buffered at protein and that
do not have significant interaction between RNA and protein.


```{r}
## Plot out effect sizes calculated from
## un-normalized data
plot_matrix <- cbind(xy.rnapro$rna, xy.rnapro$pro)

## Make a panel for 4 plots
par( mfrow = c(2,2), mar = c(4, 4, 3, 1), oma = c(1, 2, 1, 1),
     cex.main = 1.2, cex.axis = 1)
title_list <- list("RNA-Protein buffered",
                   "RNA-Protein non-buffered",
                   "RNA-Protein no interaction")
index_list <- 
  list( ii_buff_pro_buff = 
        (ii_rnaGTpro & ii_sig_rnapro) & (ii_rnaGTribo & ii_sig_rnaribo),
        ii_buff_pro_nobuff =
        (ii_proGTrna & ii_sig_rnapro) & (ii_rnaGTribo & ii_sig_rnaribo),
        ii_buff_no_interact = (!ii_sig_rnapro) & ii_rnaGTribo & ii_sig_rnaribo )
col_list <- crayon[c("Orange Red", "Pine Green", "Violet Blue")]
for (i in 1:length(index_list)) {
    plot(plot_matrix, col = "gray60", cex = .8, ann = FALSE, axes = FALSE)
    points(plot_matrix[ which( index_list[[i]] ), ], 
           col = col_list[[i]], pch = 16, cex = .7)
    abline(0, 1, v = 0, h = 0, col = crayon["Black"], lwd = .8)
    axis(1); axis(2)
    title(main = title_list[[i]],
          xlab = "log2(Human/Chimp) RPKM of RNA levels",
          ylab = "log2(Human/Chimp) of Protein levels")
}
title(main = "RNA-Ribo buffered", outer = TRUE)
```




For genes that are not buffered at the Ribo level compared to the RNA level, 
plot out per gene effect sizes of RNA and protein for genes
that are buffered at protein, non-buffered at protein and that
do not have significant interaction between RNA and protein.


```{r}
## Plot out effect sizes calculated from
## un-normalized data
plot_matrix <- cbind(xy.rnapro$rna, xy.rnapro$pro)

## Make a panel for 4 plots
par( mfrow = c(2,2), mar = c(4, 4, 3, 1), oma = c(1, 2, 1, 1),
     cex.main = 1.2, cex.axis = 1)
title_list <- list("RNA-Protein buffered",
                   "RNA-Protein non-buffered",
                   "RNA-Protein no interaction")
index_list <- 
  list( ii_nobuff_pro_buff = 
      (ii_rnaGTpro & ii_sig_rnapro) & (ii_riboGTrna & ii_sig_rnaribo),
      ii_nobuff_pro_nobuff =
      (ii_proGTrna & ii_sig_rnapro) & (ii_riboGTrna & ii_sig_rnaribo),
      ii_nobuff_no_interact = (!ii_sig_rnapro) & ii_riboGTrna & ii_sig_rnaribo )
col_list <- crayon[c("Orange Red", "Pine Green", "Violet Blue")]
for (i in 1:length(index_list)) {
    plot(plot_matrix, col = "gray60", cex = .8, ann = FALSE, axes = FALSE)
    points(plot_matrix[ which( index_list[[i]] ), ], 
           col = col_list[[i]], pch = 16, cex = .7)
    abline(0, 1, v = 0, h = 0, col = crayon["Black"], lwd = .8)
    axis(1); axis(2)
    title(main = title_list[[i]],
          xlab = "log2(Human/Chimp) RPKM of RNA levels",
          ylab = "log2(Human/Chimp) of Protein levels")
}
title(main = "RNA-Ribo non-buffered", outer = TRUE)
```


We look further into the genes that were buffered at the Ribo level
and had no significant inter-species difference between RNA and Protein.

**Question**: We see in the scatter plot, some genes have a much larger
effect size at the RNA level compared to the protein level. Yet,
these genes did not show significant difference between inter-species
differences at the RNA level and and the protein level.
    
**A possible explanation**: Maybe these genes have a larger variance at the 
RNA level compared to the protein level. Because statistial significance is 
a function of effect size and variance, perhaps the size of the variances
of these genes canceled out the large effect size of these genes.


```{r}
## Compute per gene variance estimate for RNA data and for protein data
## We first compute variances among species and then take the average of 
## the two to be the per gene variance estimates

load( file.path(rdadir,"eSetRRP.rda") )
eSet.temp <- eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]
require(matrixStats)
varmat <- lapply( seq_along( c("human","chimp") ), function(i) {
  ii <- eSet.temp$species == c("human","chimp")[i]
  eSet.tt <- eSet.temp[, ii]
  emat <- lapply( seq_along( c("ribo","rna","protein") ), function(j) {
    jj <- eSet.tt$seqData == c("ribo","rna","protein")[j]
    rowVars( exprs(eSet.tt[, jj] ), na.rm=TRUE )
  })
  emat <- do.call(cbind, emat)
  colnames(emat) <- c("ribo","rna","protein")
  return( data.frame(emat) )
})
names(varmat) <- c("human","chimp")

varmat <- data.frame(
    rna = (varmat$human$rna + varmat$chimp$rna)/2,
    pro = (varmat$human$pro + varmat$chimp$pro)/2)
```

    


```{r}
par( mfrow = c(2,2), mar = c(4, 4, 4, 1), oma = c(1, 2, 2, 1),
     cex.main = 1.2, cex.axis = 1)
ii_buff_no_interact <- (!ii_sig_rnapro) & ii_rnaGTribo & ii_sig_rnaribo 
plot(plot_matrix, col = "gray60", cex = .8, ann = FALSE, axes = FALSE)
points(plot_matrix[ which( ii_buff_no_interact ), ], 
       col = crayon["Violet Blue"], pch = 16, cex = .7)
abline(0, 1, v = 0, h = 0, col = crayon["Black"], lwd = .8)
axis(1); axis(2)
title(main = "Per gene fold changes",
      xlab = "log2(Human/Chimp) RPKM of RNA levels",
      ylab = "log2(Human/Chimp) of Protein levels")

plot(log2(varmat), col = "gray60", cex = .8, ann = FALSE, axes = FALSE)
points(log2(varmat[ which( ii_buff_no_interact ), ]), 
       col = crayon["Violet Blue"], pch = 16, cex = .7)
abline(0, 1, v = 0, h = 0, col = crayon["Black"], lwd = .8)
axis(1); axis(2)
title(main = "Per gene variance estimates",
      xlab = "log2(Human/Chimp) RPKM of RNA levels",
      ylab = "log2(Human/Chimp) of Protein levels")

title(main = "RNA-Ribo buffered genes that \n 
              has no sig. diff. between RNA and protein", outer = TRUE)


```







## Size approach

```{r}
buffering_matrix_size <- 
    matrix(0, 3, 3,
     dimnames = list(rnaribo = c("Ribo buffered", 
                                 "Ribo non-buffered",
                                 "Column sum"),
                   rnapro = c("Protein buffered",
                              "Protein non-buffered",
                              "Row sum") ) )

## Among the genes buffered at translation,
## count the number of genes buffered at protein,
## non-buffered at protein and the total number
buffering_matrix_size[1,1] <- 
    sum( (ii_rnaGTpro)[ii_rnaGTribo & ii_sig_rnaribo])
buffering_matrix_size[1,2] <- 
    sum( (ii_proGTrna)[ii_rnaGTribo & ii_sig_rnaribo])
buffering_matrix_size[1,3] <- 
    sum(ii_rnaGTribo & ii_sig_rnaribo)


## Among the genes non-buffered at translation,
## count the number of genes buffered at protein,
## non-buffered at protein and the total number
buffering_matrix_size[2,1] <- 
    sum( (ii_rnaGTpro )[ii_riboGTrna & ii_sig_rnaribo])
buffering_matrix_size[2,2] <- 
    sum( (ii_proGTrna )[ii_riboGTrna & ii_sig_rnaribo])
buffering_matrix_size[2,3] <- 
  sum(ii_riboGTrna & ii_sig_rnaribo)


## Of all genes with sig. difference in translation efficiency,
## count the number of genes buffered at protein,
## non-buffered at protein, and no interaction between RNA and protein
buffering_matrix_size[3,1] <- 
    sum( (ii_rnaGTpro)[ 
        (ii_rnaGTribo | ii_riboGTrna) & ii_sig_rnaribo] )
buffering_matrix_size[3,2] <- 
    sum( (ii_proGTrna)[ 
        (ii_rnaGTribo | ii_riboGTrna) & ii_sig_rnaribo ] )

## No. of genes with sig. translation efficiency
buffering_matrix_size[3,3] <- sum(ii_sig_rnaribo)

kable(buffering_matrix_size)
```



For genes buffered at the Ribo level compared to the RNA level, 
plot out per gene effect sizes of RNA and protein for genes
that are buffered at protein and  non-buffered at protein.


```{r}
## Plot out effect sizes calculated from
## un-normalized data
plot_matrix <- cbind(xy.rnapro$rna, xy.rnapro$pro)

## Make a panel for 4 plots
par( mfrow = c(2,2), mar = c(4, 4, 3, 1), oma = c(1, 2, 1, 1),
     cex.main = 1.2, cex.axis = 1)
title_list <- list("RNA-Protein buffered",
                   "RNA-Protein non-buffered",
                   "RNA-Protein no interaction")
index_list <- 
  list( ii_buff_pro_buff = 
        (ii_rnaGTpro) & (ii_rnaGTribo & ii_sig_rnaribo),
        ii_buff_pro_nobuff =
        (ii_proGTrna) & (ii_rnaGTribo & ii_sig_rnaribo) )
col_list <- crayon[c("Orange Red", "Pine Green")]
for (i in 1:length(index_list)) {
    plot(plot_matrix, col = "gray60", cex = .8, ann = FALSE, axes = FALSE)
    points(plot_matrix[ which( index_list[[i]] ), ], 
           col = col_list[[i]], pch = 16, cex = .7)
    abline(0, 1, v = 0, h = 0, col = crayon["Black"], lwd = .8)
    axis(1); axis(2)
    title(main = title_list[[i]],
          xlab = "log2(Human/Chimp) RPKM of RNA levels",
          ylab = "log2(Human/Chimp) of Protein levels")
}
title(main = "RNA-Ribo buffered", outer = TRUE)
```




For genes that are not buffered at the Ribo level compared to the RNA level, 
plot out per gene effect sizes of RNA and protein for genes
that are buffered at protein, non-buffered at protein.


```{r}
## Plot out effect sizes calculated from
## un-normalized data
plot_matrix <- cbind(xy.rnapro$rna, xy.rnapro$pro)

## Make a panel for 4 plots
par( mfrow = c(2,2), mar = c(4, 4, 3, 1), oma = c(1, 2, 1, 1),
     cex.main = 1.2, cex.axis = 1)
title_list <- list("RNA-Protein buffered",
                   "RNA-Protein non-buffered",
                   "RNA-Protein no interaction")
index_list <- 
  list( ii_nobuff_pro_buff = 
      (ii_rnaGTpro) & (ii_riboGTrna & ii_sig_rnaribo),
      ii_nobuff_pro_nobuff =
      (ii_proGTrna) & (ii_riboGTrna & ii_sig_rnaribo) )
col_list <- crayon[c("Orange Red", "Pine Green")]
for (i in 1:length(index_list)) {
    plot(plot_matrix, col = "gray60", cex = .8, ann = FALSE, axes = FALSE)
    points(plot_matrix[ which( index_list[[i]] ), ], 
           col = col_list[[i]], pch = 16, cex = .7)
    abline(0, 1, v = 0, h = 0, col = crayon["Black"], lwd = .8)
    axis(1); axis(2)
    title(main = title_list[[i]],
          xlab = "log2(Human/Chimp) RPKM of RNA levels",
          ylab = "log2(Human/Chimp) of Protein levels")
}
title(main = "RNA-Ribo non-buffered", outer = TRUE)
```



Session information
===================

```{r}
sessionInfo()
```



