---
output:
  knitrBootstrap::bootstrap_document:
    title: "RNA-Ribo divergence conditioned on RNA-Protein divergence"
    theme: default
---


RNA-Ribo divergence conditioned on RNA-Protein divergence
===========================================================


```{r run_author, results='asis', echo=FALSE}
email <- "<a href='mailto:joyce.hsiao1@gmail.com'>Joyce Hsiao</a>"
cat(paste("Author:", email))
```


```{r run_date, results='asis', echo=FALSE}
#last_update = format(Sys.time(), "(<time>%Y-%m-%d</time>)")
last_update <- Sys.time()
cat(paste("Last updated:", last_update))
```


```{r knitr_settings, include=FALSE, echo=FALSE}
library(knitr)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(message = FALSE, warning = FALSE, eval = TRUE, 
               echo = TRUE,
               bootstrap.panel = TRUE, bootstrap.show.code = TRUE)
```


```{r load_packages}
library(Biobase)

## Set working directories 
dir <- "../"
figdir <- file.path(dir,"figures")
rdadir <- file.path(dir,"rdas")
datadir <- file.path(dir, "data")

## Use broman package for colors
require(broman)
crayon <- brocolors("crayons")

## Load customized packages
require(devtools)
require(Humanzee)
```


Goal
====

We investigate protein divergence relative to RNA divergence and consider
if attenuation at the protein level is more likely to be associated with
attenuation at the RNA level, and if enhancement at the protein level 
is more likely to be associated with enhancement at the RNA level.


Definitions
===========

Of all the 3,188 genes included in the analysis, we apply divergence
definition version-20150812 and investigate genes that fall into the following 
three categories

1. **attenuated** at the protein level relative to the RNA level 
    * Sig. different between RNA and protein divergence) *AND* 
      ( (the absolute value of protein divergence 
      is smaller than the absolute value of RNA divergence) *OR* (protein 
      divergence is in the opposite direction of RNA divergence) ) 

2. **enhanced** at the protein level relative to the RNA level
    * Sig. different between RNA and protein divergence) *AND* 
      ( (the absolute value of protein divergence is greater 
      than the absolute value of RNA divergence) *AND* 
      (protein divergence is in the opposite direction of RNA divergence) )

3. **null divergence**: of the 3,188 genes, those that do not fall in the 
    above two categories.
    


```{r}
## Load rdas of LRT results and compute fold change estimates using
## RPKM values before normalization. 

## load LRT results of RNA vs. Ribo divergence
load(file.path(rdadir,"TEnew.rda"))

## load LRT results of RNA vs. Protein divergence
load(file.path(rdadir,"rnapro.rda"))


## Compute fold changes based on un-normalized data
## from the object eSetRRP.log2 in eSetRRP.rda
load( file.path(rdadir,"eSetRRP.rda") )

eSet.temp <- eSetRRP.log2[ ,eSetRRP.log2$species!="rhesus"]

fc.mat <- lapply( seq_along( c("human","chimp") ), function(i) {
  ii <- eSet.temp$species == c("human","chimp")[i]
  eSet.tt <- eSet.temp[, ii]
  emat <- lapply( seq_along( c("ribo","rna","protein") ), function(j) {
    jj <- eSet.tt$seqData == c("ribo","rna","protein")[j]
    rowMeans( exprs(eSet.tt[, jj] ), na.rm=TRUE )
  })
  emat <- do.call(cbind, emat)
  colnames(emat) <- c("ribo","rna","protein")
  return( data.frame(emat) )
})
names(fc.mat) <- c("human","chimp")

dmat_unnormed <- data.frame( ribo = fc.mat$human$ribo-fc.mat$chimp$ribo,
                             rna = fc.mat$human$rna-fc.mat$chimp$rna,
                             pro = fc.mat$human$protein-fc.mat$chimp$protein )

xy.rnapro <- data.frame( rna = dmat_unnormed$rna,
                         pro = dmat_unnormed$pro )
xy.riborna <- data.frame( rna = dmat_unnormed$rna,
                          ribo = dmat_unnormed$ribo )
xy.ribopro <- data.frame( ribo = dmat_unnormed$ribo,
                          pro = dmat_unnormed$pro )
```


Gene classification
===================

## RNA vs. Ribo divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and Ribo
ii_sig_rnaribo <- res.riborna$int.qval < .01


## Category 1 - Enhanced divergence, criteria are as follows
## a) Sig. RNA vs. Ribo difference, AND
## b) ( abs(RNA FC) < abs(Ribo FC) ) AND ( RNA FC * Ribo FC > 0)
ii_riboGTrna <- (ii_sig_rnaribo) & 
                ( (abs(xy.riborna$ribo) > abs(xy.riborna$rna)) & 
                  (xy.riborna$rna * xy.riborna$ribo > 0 ) )


## Category 2 - Attenuated divergence, criteria are as follows
## a) Sig. RNA vs. Ribo diff., AND
## b) ( abs(RNA FC) > abs(Ribo FC) ) AND ( RNA FC * Ribo FC < 0)
ii_rnaGTribo <- (ii_sig_rnaribo) &  
                ( (abs(xy.riborna$rna) > abs(xy.riborna$ribo) ) | 
                  (xy.riborna$rna * xy.riborna$ribo < 0) ) 


Ribo_refmat <- rbind.data.frame(attenuated = sum(ii_rnaGTribo),
                                   enhanced = sum(ii_riboGTrna) )
colnames(Ribo_refmat) <- "# genes"
kable(Ribo_refmat)
```


## RNA vs. Protein divergence

```{r}
## Genes with sig. human/chimp difference in divergence between RNA and protein
ii_sig_rnapro <- res.rnapro$int.qval < .01

## Category 1 - Enhanced divergence, criteria are as follows
## a) Sig. RNA vs. Protein difference, AND
## b) ( abs(RNA FC) < abs(Protein FC) ) AND ( RNA FC * Protein FC > 0)
ii_proGTrna <- (ii_sig_rnapro) &  
               ( (abs(xy.rnapro$pro) > abs(xy.rnapro$rna)) & 
                 (xy.rnapro$rna * xy.rnapro$pro > 0 ) )


## Category 2 - Attenuated divergence, criteria are as follows
## a) Sig. RNA vs. Protein diff., AND
## b) ( abs(RNA FC) > abs(Protein FC) ) OR ( RNA FC * Protein FC < 0)
ii_rnaGTpro <- (ii_sig_rnapro) &  
               ( (abs(xy.rnapro$rna) > abs(xy.rnapro$pro) ) | 
                 (xy.rnapro$rna * xy.rnapro$pro < 0) ) 


Protein_refmat <- rbind.data.frame( attenuated = sum(ii_rnaGTpro),
                                    enhanced = sum(ii_proGTrna) )
colnames(Protein_refmat) <- "# genes"
kable(Protein_refmat)
```




RNA ~ Protein
==============


Among ``r sum( !(ii_rnaGTpro | ii_proGTrna) )`` genes that are neither 
enhanced nor attenuated at the protein level compared to the RNA level 
(RNA ~ Protein), 

* ``r sum( !(ii_rnaGTpro | ii_proGTrna) & ii_rnaGTribo)`` genes are 
  attenuated at the translation level relative to the transcript level, and

* ``r sum( !(ii_rnaGTpro | ii_proGTrna) & ii_riboGTrna)`` genes are


```{r}
ii_null_rnapro <- !(ii_rnaGTpro | ii_proGTrna)
ii_null_rnapro_riboAttenuated <- !(ii_rnaGTpro | ii_proGTrna) & ii_rnaGTribo
ii_null_rnapro_riboEnhanced <- !(ii_rnaGTpro | ii_proGTrna) & ii_riboGTrna

boxplot(abs(dmat_unnormed$rna[ii_null_rnapro]),
        abs(dmat_unnormed$ribo[ii_null_rnapro]),
        abs(dmat_unnormed$pro[ii_null_rnapro]), 
        axes = F, ylim = c(0, 10))
axis(1, at = c(1,2,3), labels = c("RNA", "Ribo", "Protein"), 
     col = "white")
axis(2)
points_cols <- c("Scarlet", "Sunglow", "Shamrock")
points(rep(1, sum(ii_null_rnapro)), 
       abs(dmat_unnormed$rna[ii_null_rnapro]), 
       pch = 19, cex = .5, 
       col = crayon[points_cols[1]] )
points(rep(2, sum(ii_null_rnapro)), 
       abs(dmat_unnormed$ribo[ii_null_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[2]] )
points(rep(3, sum(ii_null_rnapro)), 
       abs(dmat_unnormed$pro[ii_null_rnapro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[3]] )
text(2, 10, adj = 0,
     label = paste("RNA ~ Protein", sum(ii_null_rnapro), "genes" ) )
text(2, 9, adj = 0,
     label = paste("RNA vs. Ribo sig. diff.", 
       sum(ii_null_rnapro & (res.riborna$int.qval<.01)), "genes; \n",
       round(100*sum(ii_null_rnapro & (res.riborna$int.qval<.01))/
               sum(ii_null_rnapro), 2), "%") )
text(2, 8, adj = 0,
     label = paste("Attenuated translation", 
       sum(ii_null_rnapro_riboAttenuated), "genes; \n",
       round(100*sum(ii_null_rnapro_riboAttenuated)/sum(ii_null_rnapro), 2), "%") )
text(2, 7, adj = 0,
     label = paste("Enhanced translation", 
       sum(ii_null_rnapro_riboEnhanced), "genes; \n",
       round(100*sum(ii_null_rnapro_riboEnhanced)/sum(ii_null_rnapro), 2), "%") )
title(main = "RNA ~ Protein, no sig. diff. divergence",
      ylab = "absolute divergence (human-chimpanzee fold change)")

# # Compute pairwise p-values between effect sizes of Ribo, RNA, and protein
# # among genes with RNA ~ Protein
# rna_ribo_t <- t.test( abs(dmat_unnormed$rna[ii_proNULLrna]),
#        abs(dmat_unnormed$ribo[ii_proNULLrna]), paired = TRUE )
# rna_pro_t <- t.test( abs(dmat_unnormed$rna[ii_proNULLrna]),
#        abs(dmat_unnormed$pro[ii_proNULLrna]), paired = TRUE )
# ribo_pro_t <- t.test( abs(dmat_unnormed$ribo[ii_proNULLrna]),
#        abs(dmat_unnormed$pro[ii_proNULLrna]), paired = TRUE )
# 
# text(.5, 10, adj = 0,
#      label = paste("Paired t-test of effect sizes") )
# text(.5, 9, adj = 0,
#      label = paste("RNA vs. Ribo: p-value < 1e-14") )
# text(.5, 8, adj = 0,
#      label = paste("RNA vs. Protein: p-value < 1e-15") )
# text(.5, 7, adj = 0,
#      label = paste("Ribo vs. Protein: p-value < 1e-15") )
# 
```


## Signal in RNA ~ Protein

Among the ``r sum(ii_null_rnapro & res.riborna$int.qval < .01)`` genes with 
RNA and Ribo significant differences and no significant difference between 
RNA and protein, we do a follow-up analysis to look further to consider the contribution/relation of translation divergence to protein divergence, 
relative to RNA divergence. 

Yoav: So, for example, if gene 1 (among the 
``r sum(ii_null_rnapro & res.riborna$int.qval < .01)`` genes with sig. diff. 
between RNA and Ribo) shows attenuated translation relative to RNA, protein 
divergence will be attenuated relative to RNA divergence. If gene 2 
(among the ``r sum(ii_null_rnapro & res.riborna$int.qval < .01)`` genes with 
sig. diff. between RNA and Ribo) shows enhanced ribo divergence compared with 
RNA, I predict that the protein divergence will also be enhanced compared 
with RNA.


```{r}
## RNA-Ribo enhancement or attenuation
ii_null_rnapro_riboAttenuated <- !(ii_rnaGTpro | ii_proGTrna) & ii_rnaGTribo
ii_null_rnapro_riboEnhanced <- !(ii_rnaGTpro | ii_proGTrna) & ii_riboGTrna

## RNA-Protein enhancement or attenuation
ii_null_rnapro_proAttenuated <- 
    !(ii_rnaGTpro | ii_proGTrna) & 
    ( abs(xy.rnapro$rna) > abs(xy.rnapro$pro) | 
      xy.rnapro$rna * xy.rnapro$pro < 0 ) 

ii_null_rnapro_proEnhanced <- 
    !(ii_rnaGTpro | ii_proGTrna) &
    ( abs(xy.rnapro$rna) < abs(xy.rnapro$pro) & 
      (xy.rnapro$rna * xy.rnapro$pro > 0)  )

plot_mat <- matrix(0, 2, 2)
rownames(plot_mat) <- c("RNA-Ribo enhanced", "RNA-Ribo attenuated")
colnames(plot_mat) <- c("RNA-Protein enhanced", "RNA-Protein attenuated")

plot_mat[1,1] <- sum(ii_null_rnapro_riboEnhanced*ii_null_rnapro_proEnhanced)
plot_mat[1,2] <- sum(ii_null_rnapro_riboEnhanced*ii_null_rnapro_proAttenuated)
plot_mat[2,1] <- sum(ii_null_rnapro_riboAttenuated*ii_null_rnapro_proEnhanced)
plot_mat[2,2] <- sum(ii_null_rnapro_riboAttenuated*ii_null_rnapro_proAttenuated)

kable(plot_mat)
```


### Attenuated translation

Among the ``r sum(ii_null_rnapro_riboAttenuated)`` genes that were attenuated 
at the translation level relative to the transcript level 
(RNA-Ribo attenuation), we tested the null hypothesis that these genes are 
as likely to be attenuated as to be enhanced at the protein level relative to 
the transcript level (RNA-Protein). If this null hypothesis were rejected and 
the number of genes that were attenuated at the protein layer is larger than 
the number of genes that were enhanced at the protein layer, then we say that 
attenuation at the translation level is more likey to co-occur with
attenuation at the protein level than with enhancement at the protein level.


```{r}
prop.test(plot_mat[2,2], rowSums(plot_mat)[2])
```


### Enhanced translation

Among the ``r sum(ii_null_rnapro_riboEnhanced)`` genes that were reinforced 
at translation, we tested the null hypothesis that these genes are as likely
to be enhanced as to be attenuated at the protein layer. If this null 
hypothesis were rejected and the number of genes that were enhanced at the 
protein layer is more than those attenuated at the protein layer, then we 
say that genes that are enhanced at the translation level are more likely 
to be enhanced at the protein level than to be attenuated at the protein level.


```{r}
## One-proportion z-test
prop.test(plot_mat[1,1], rowSums(plot_mat)[1])

## Exact test of proportions (binomial test)
binom.test(plot_mat[1,1], rowSums(plot_mat)[1])
```





RNA > Protein
==============

Among the genes that are attenuated at the protein level compared to the 
RNA level (RNA > protein), we are interested in the number of genes with
attenuated translation. In other words, we ask whether genes that were
attenuated at the protein level were more likely to be enriched with
genes attenuated at the translation level.

For the hypothesis testing, we computed number of genes enriched in 
translation
attenuation among the genes enriched for protein attenuation in the
set of ``r sum(ii_null_rnapro)`` with similar divergence levels between
RNA and protein.


```{r}
boxplot(abs(dmat_unnormed$rna[ii_rnaGTpro]),
        abs(dmat_unnormed$ribo[ii_rnaGTpro]),
        abs(dmat_unnormed$pro[ii_rnaGTpro]), 
        axes = F, ylim = c(0, 10))
axis(1, at = c(1,2,3), labels = c("RNA", "Ribo", "Protein"), 
     col = "white")
axis(2)
points_cols <- c("Scarlet", "Sunglow", "Shamrock")
points(rep(1, sum(ii_rnaGTpro)), abs(dmat_unnormed$rna[ii_rnaGTpro]), 
       pch = 19, cex = .5, 
       col = crayon[points_cols[1]] )
points(rep(2, sum(ii_rnaGTpro)), abs(dmat_unnormed$ribo[ii_rnaGTpro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[2]] )
points(rep(3, sum(ii_rnaGTpro)), abs(dmat_unnormed$pro[ii_rnaGTpro]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[3]] )
text(2.2, 9, adj = 0,
     label = paste("RNA > Protein", sum(ii_rnaGTpro), 
                   "genes" ) )
text(2.2, 8, adj = 0,
     label = paste("RNA vs. Ribo sig. diff.", 
      sum( ii_rnaGTpro*(res.riborna$int.qval<.01) ), 
      "genes; \n",
      round( 100*sum(ii_rnaGTpro*(res.riborna$int.qval<.01) )/
            sum(ii_rnaGTpro), 2),"%") )
text(2.2, 7, adj = 0,
     label = paste("Attenuated at translation", 
      sum( ii_rnaGTpro*ii_rnaGTribo ), "genes; \n",
      round( 100*sum(ii_rnaGTpro*ii_rnaGTribo)/sum(ii_rnaGTpro), 2),"%") )
title(main = paste("RNA > Protein",sum(ii_rnaGTpro),
                   "genes (Attenuated divergence)"),
      ylab = "absolute divergence (human-chimpanzee fold change)")


## Null proportion of RNA-Protein attenuated | RNA-Ribo attenuated
rnapro_null_proAttenuated <- plot_mat[2, 2]/rowSums(plot_mat)[2]


prop.test( sum(ii_rnaGTpro*ii_rnaGTribo),
           sum(ii_rnaGTpro), p = rnapro_null_proAttenuated)

binom.test( sum(ii_rnaGTpro*ii_rnaGTribo),
           sum(ii_rnaGTpro), p = rnapro_null_proAttenuated)
```





RNA < Protein
==============

Among the genes that are enhanced at the protein level relative to the RNA 
level (RNA < Protein), we are interested in the number of genes that are 
enhanced at the translation level relative to the RNA level.

For the hypothesis testing, we computed number of genes enriched in 
translation reinforcement among the genes enriched for protein 
enhancement in the
set of ``r sum(ii_null_rnapro)`` with similar divergence levels between
RNA and protein.



```{r}
boxplot(abs(dmat_unnormed$rna[ii_proGTrna]),
        abs(dmat_unnormed$ribo[ii_proGTrna]),
        abs(dmat_unnormed$pro[ii_proGTrna]), 
        axes = F, ylim = c(0, 10))
axis(1, at = c(1,2,3), labels = c("RNA", "Ribo", "Protein"), 
     col = "white")
axis(2)
points_cols <- c("Scarlet", "Sunglow", "Shamrock")
points(rep(1, sum(ii_proGTrna)), abs(dmat_unnormed$rna[ii_proGTrna]), 
       pch = 19, cex = .5, 
       col = crayon[points_cols[1]] )
points(rep(2, sum(ii_proGTrna)), abs(dmat_unnormed$ribo[ii_proGTrna]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[2]] )
points(rep(3, sum(ii_proGTrna)), abs(dmat_unnormed$pro[ii_proGTrna]), 
       pch = 19, cex = .5,
       col = crayon[points_cols[3]] )
text(1.5, 9, adj = 0,
     label = paste("Protein > RNA", sum(ii_proGTrna), "genes" ) )
text(1.5, 8, adj = 0,
     label = paste("RNA vs. Ribo sig. diff.", 
       sum( ii_proGTrna*(res.riborna$int.qval<.01) ), "genes; \n",
       round(100*sum( ii_proGTrna*(res.riborna$int.qval<.01) )/
             sum(ii_proGTrna), 2), "%") ) 
text(1.5, 6, adj = 0,
     label = paste("Reinfored at translation", 
      sum(ii_proGTrna*ii_riboGTrna), "genes; \n",
      round(100*sum(ii_proGTrna*ii_riboGTrna)/sum(ii_proGTrna),
            2),"%") )
title(main = paste("RNA < Protein",sum(ii_proGTrna),
                  "genes (Enhanced protein divergence)"),
      ylab = "absolute divergence (human-chimpanzee fold change)")


## Null proportion of RNA-Ribo enhanced | RNA-Protein enhanced
rnapro_null_proEnhanced <- plot_mat[1, 1]/rowSums(plot_mat)[1]

prop.test( sum(ii_proGTrna*ii_riboGTrna),
           sum(ii_proGTrna), p = rnapro_null_proEnhanced)

binom.test( sum(ii_proGTrna*ii_riboGTrna),
           sum(ii_proGTrna), p = rnapro_null_proEnhanced)
```



