---
title: "include_additional_human_data_for_batch_effect"
author: "Sidney Wang"
date: "November 3, 2015"
output: html_document
---

Additional data from mm2 and mm7 were processed and included in "HCR.plus.ribo" data object to properly model batch effect and to further correct it.

load data
```{r}
load("../rdas/HRC.plus.ribo.rda")
load("../rdas/mm4.ribo.rda")
load("../rdas/HCR.ortho.geneLength.rda")
```

compute RPKM
=============
use species specific gene length

```{r}
HCR.plus.ribo.rpkm <- prop.table(as.matrix(HCR.plus.ribo[,-1]),2)*10^9
HCR.plus.ribo.rpkm <- cbind(HCR.plus.ribo.rpkm[,1:5]/HCR.geneLength$Chimp,HCR.plus.ribo.rpkm[,6:11]/HCR.geneLength$Human,HCR.plus.ribo.rpkm[,12:16]/HCR.geneLength$Rhesus,HCR.plus.ribo.rpkm[,17:26]/HCR.geneLength$Human)


mm4.ribo.rpkm <- prop.table(as.matrix(mm4.ribo[,-1]),2)*10^9
mm4.ribo.rpkm<-cbind(mm4.ribo.rpkm[,1:4]/HCR.geneLength$Chimp, mm4.ribo.rpkm[,5:8]/HCR.geneLength$Rhesus, mm4.ribo.rpkm[,9:12]/HCR.geneLength$Human)

```

filter data by requiring at 2/3 (8) of mm4 data have rpkm greater than 0

```{r}
rpkm.filter<-c() 
for (i in 1:length(mm4.ribo.rpkm[,1])){
rpkm.filter[i]<-sum(mm4.ribo.rpkm[i,] == 0) < 5
}

HCR.plus.ribo.rpkm<-HCR.plus.ribo.rpkm[rpkm.filter,]
mm4.ribo.rpkm<-mm4.ribo.rpkm[rpkm.filter,]
```

quantile normalize and log2 transform data
```{r}
library(limma)
HCR.plus.ribo.rpkm.Q<-normalizeQuantiles(HCR.plus.ribo.rpkm,ties = T)
mm4.ribo.rpkm.Q<-normalizeQuantiles(mm4.ribo.rpkm,ties = T)

is.na(HCR.plus.ribo.rpkm.Q)<-which(HCR.plus.ribo.rpkm.Q == 0)
log2.HCR.plus.ribo.rpkm.Q<-log2(HCR.plus.ribo.rpkm.Q)

is.na(mm4.ribo.rpkm.Q)<-which(mm4.ribo.rpkm.Q == 0)
log2.mm4.ribo.rpkm.Q<-log2(mm4.ribo.rpkm.Q)
```


limma remove batch effect

```{r}

spec<-substring(colnames(log2.HCR.plus.ribo.rpkm.Q),1,1)
spec[17:26]<-rep("H",10)
spec.design <- model.matrix(~spec)
library(gdata)
read.xls("/Users/siddisis/Dropbox/work/Ribosome_profiling/Primate_comparison/primate_ribo_gender_mmbatch.xlsx")->cell.source
cell.source<-cell.source[,1:4]
batch.additional<-read.xls("/Users/siddisis/Dropbox/work/Ribosome_profiling/Primate_comparison/mmbatch_info_additional_YRI_for_batch_correction.xlsx")

batch<-c(as.character(cell.source$libmix),c(as.character(batch.additional$libmix[4:7]),as.character(batch.additional$libmix[-4:-7])))

batch.removed.limma.full<-removeBatchEffect(log2.HCR.plus.ribo.rpkm.Q,batch = as.factor(batch),design = spec.design)

```

limma, weighted least square
```{r}
var.weight<-matrix(nrow =length(log2.HCR.plus.ribo.rpkm.Q[,1]),ncol = 26 )

for (i in 1:length(log2.HCR.plus.ribo.rpkm.Q[,1])){
#chimp.var<-var(log2.HCR.plus.ribo.rpkm.Q[i,1:5],na.rm = T)
mm1.var<-var(log2.HCR.plus.ribo.rpkm.Q[i,6:9],na.rm = T)
mm2.var<-var(log2.HCR.plus.ribo.rpkm.Q[i,c(10,21,22,23)],na.rm = T)
mm7.var<-var(log2.HCR.plus.ribo.rpkm.Q[i,c(11,24,25,26)],na.rm = T)
#rhesus.var<-var(log2.HCR.plus.ribo.rpkm.Q[i,12:16],na.rm = T)
mm4.var<-var(log2.HCR.plus.ribo.rpkm.Q[i,c(1:3,5,13:20)],na.rm = T)
mm8.var<-var(log2.HCR.plus.ribo.rpkm.Q[i,c(4,12)],na.rm = T)
var.weight[i,]<-c(rep(1/mm4.var,3),1/mm8.var,1/mm4.var,rep(1/mm1.var,4),1/mm2.var,1/mm7.var,1/mm8.var,rep(1/mm4.var,8),rep(1/mm2.var,3),rep(1/mm7.var,3))
}

batch.removed.limma.weighted<-removeBatchEffect(log2.HCR.plus.ribo.rpkm.Q,batch = as.factor(batch),design = spec.design,weight=var.weight)

```



Combat
```{r, cache=TRUE}
library(sva)


batch.removed.combat.par<-ComBat(dat = log2.HCR.plus.ribo.rpkm.Q,mod = model.matrix(~ spec),batch = as.factor(batch),par.prior = TRUE,prior.plots = FALSE)

batch.removed.combat.nonpar<-ComBat(dat = log2.HCR.plus.ribo.rpkm.Q,mod = model.matrix(~ spec),batch = as.factor(batch),par.prior = FALSE,prior.plots = FALSE)



```



study design
```{r}
table(batch,spec)
```



pca
```{r}
library(ggplot2)
library(reshape2)
library(dplyr)
library(gdata)

# full data
pc<-prcomp(t(na.omit(log2.HCR.plus.ribo.rpkm.Q)),center = T,scale = T)
barplot((pc$sdev)^2/sum((pc$sdev)^2), names.arg =  colnames(pc$x))

temp.data<-as.data.frame(cbind(melt(pc$x),pc$x[,1],spec,batch))
names(temp.data)<-c("ID","PC","value","pc1","species","batch")

temp.data %>% filter(as.numeric(temp.data$PC) < 10) %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = species, shape = batch))+facet_wrap(~PC)+labs(y="pcX")

temp.data %>% filter(as.numeric(temp.data$PC) < 10) %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(shape = species))+facet_wrap(~PC)+labs(y="pcX")
temp.data %>% filter(as.numeric(temp.data$PC) < 10) %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(shape = batch))+facet_wrap(~PC)+labs(y="pcX")

# mm4
pc<-prcomp(t(na.omit(log2.mm4.ribo.rpkm.Q)),center = T,scale = T)
barplot((pc$sdev)^2/sum((pc$sdev)^2), names.arg =  colnames(pc$x))
mm4.spec<-substring(rownames(pc$x),1,1)
mm4.spec<-c(mm4.spec[1:8],rep("H",4))
temp.data<-as.data.frame(cbind(melt(pc$x),pc$x[,1],mm4.spec))
names(temp.data)<-c("ID","PC","value","pc1","species")

temp.data %>% filter(as.numeric(temp.data$PC) < 10) %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(shape = species))+facet_wrap(~PC)+labs(y="pcX")

# combat nonparametric corrected all 26
pc<-prcomp(t(na.omit(batch.removed.combat.nonpar)),center = T,scale = T)
barplot((pc$sdev)^2/sum((pc$sdev)^2), names.arg =  colnames(pc$x))
temp.data<-as.data.frame(cbind(melt(pc$x),pc$x[,1],spec,batch))
names(temp.data)<-c("ID","PC","value","pc1","species","batch")

temp.data %>% filter(as.numeric(temp.data$PC) < 10) %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = species, shape=batch))+facet_wrap(~PC)

# combat nonparametric corrected, 16 used 
pc<-prcomp(t(na.omit(batch.removed.combat.nonpar[,1:16])),center = T,scale = T)
barplot((pc$sdev)^2/sum((pc$sdev)^2), names.arg =  colnames(pc$x))
temp.data<-as.data.frame(cbind(melt(pc$x),pc$x[,1],spec[1:16],batch[1:16]))
names(temp.data)<-c("ID","PC","value","pc1","species","batch")

temp.data %>% filter(as.numeric(temp.data$PC) < 10) %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(color = species, shape=batch))+facet_wrap(~PC)+labs(y="PCX")

temp.data %>% filter(as.numeric(temp.data$PC) < 10) %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(shape = species))+facet_wrap(~PC)+labs(y="pcX")
temp.data %>% filter(as.numeric(temp.data$PC) < 10) %>% ggplot(aes(x=pc1,y=value))+geom_point(aes(shape = batch))+facet_wrap(~PC)+labs(y="pcX")



## pc1 vs pc2

pc.mm4<-prcomp(t(na.omit(log2.mm4.ribo.rpkm.Q)),center = T,scale = T)
pc1.mm4<-pc.mm4$x[,1]
pc2.mm4<-pc.mm4$x[,2]

mm4.pc.data<-as.data.frame(cbind(pc1.mm4,pc2.mm4))
mm4.pc.data<-as.data.frame(cbind(rownames(mm4.pc.data),rep("mm4",12),mm4.pc.data,c(rep("C",4),rep("R",4),rep("H",4)),rep("mm4",12)),stringsAsFactors=F)
names(mm4.pc.data)<-c("ID","data","PC1","PC2","species","batch")

pc.ori<-prcomp(t(na.omit(log2.HCR.plus.ribo.rpkm.Q[,1:16])),center = T,scale = T)
pc1.ori<-pc.ori$x[,1]
pc2.ori<-pc.ori$x[,2]

pc.limma<-prcomp(t(na.omit(batch.removed.limma.full[,1:16])),center = T,scale = T)
pc1.limma<-pc.limma$x[,1]
pc2.limma<-pc.limma$x[,2]

pc.limma.weighted<-prcomp(t(na.omit(batch.removed.limma.weighted[,1:16])),center = T,scale = T)
pc1.limma.wt<-pc.limma.weighted$x[,1]
pc2.limma.wt<-pc.limma.weighted$x[,2]

pc.combat.par<-prcomp(t(na.omit(batch.removed.combat.par[,1:16])),center = T,scale = T)
pc1.combat.par<-pc.combat.par$x[,1]
pc2.combat.par<-pc.combat.par$x[,2]

pc.combat.nonpar<-prcomp(t(na.omit(batch.removed.combat.nonpar[,1:16])),center = T,scale = T)
pc1.combat.nonpar<-pc.combat.nonpar$x[,1]
pc2.combat.nonpar<-pc.combat.nonpar$x[,2]

pc1.data<-cbind(pc1.ori,pc1.limma,pc1.limma.wt,pc1.combat.par,pc1.combat.nonpar)
colnames(pc1.data)<-c("original","limma","limma.wt","combat.par","combat.nonpar")
pc1.data<-melt(pc1.data)
pc2.data<-melt(cbind(pc2.ori,pc2.limma,pc2.limma.wt,pc2.combat.par,pc2.combat.nonpar))

pc.data<-cbind(pc1.data,pc2.data$value,spec[1:16],batch[1:16])
names(pc.data)<-c("ID","data","PC1","PC2","species","batch")

#pc.data %>% ggplot(aes(x=PC1,y=PC2))+geom_point(aes(color = species, shape=batch))+facet_wrap(~data)+theme(panel.grid.minor=element_blank())

rbind(mm4.pc.data,pc.data)%>% ggplot(aes(x=PC1,y=PC2))+geom_point(aes(color = species, shape=batch))+facet_wrap(~data)+theme(panel.grid.minor=element_blank())

rbind(mm4.pc.data,pc.data)%>% ggplot(aes(x=PC1,y=PC2))+geom_point(aes( shape=species))+facet_wrap(~data)+theme(panel.grid.minor=element_blank())

rbind(mm4.pc.data,pc.data)%>% ggplot(aes(x=PC1,y=PC2))+geom_point(aes( shape=batch))+facet_wrap(~data)+theme(panel.grid.minor=element_blank())



```



DE test beta without batch correction

```{r}
log2.HCR.ribo.rpkm.Q<-log2.HCR.plus.ribo.rpkm.Q[,1:16]

species.label<-as.factor(substring(colnames(log2.HCR.ribo.rpkm.Q),1,1))

ori.lm.fit<-lmFit(log2.HCR.ribo.rpkm.Q,design = model.matrix(~ species.label))
oriHvC.effect<-ori.lm.fit$coefficient[,2]
oriRvC.effect<-ori.lm.fit$coefficient[,3]

ori.de.test<-eBayes(ori.lm.fit)

oriHvC.pval<-ori.de.test$p.value[,2]
oriRvC.pval<-ori.de.test$p.value[,3]

```


DE test, beta for mm4

```{r}
species.label<-as.factor(substring(colnames(log2.mm4.ribo.rpkm.Q),1,1))

mm4.lm.fit<-lmFit(log2.mm4.ribo.rpkm.Q,design = model.matrix(~ species.label))
mm4HvC.effect<-mm4.lm.fit$coefficient[,3]
mm4RvC.effect<-mm4.lm.fit$coefficient[,2]

mm4.de.test<-eBayes(mm4.lm.fit)

mm4HvC.pval<-mm4.de.test$p.value[,3]
mm4RvC.pval<-mm4.de.test$p.value[,2]

```



DE test beta for batch corrected set

```{r}

species.label<-as.factor(substring(colnames(log2.HCR.ribo.rpkm.Q),1,1))

# limma

HCR.ribo.batch.corrected<-batch.removed.limma.full[,1:16]

limma.lm.fit<-lmFit(HCR.ribo.batch.corrected,design = model.matrix(~ species.label))
limmaHvC.effect<-limma.lm.fit$coefficient[,2]
limmaRvC.effect<-limma.lm.fit$coefficient[,3]

limma.de.test<-eBayes(limma.lm.fit)

limmaHvC.pval<-limma.de.test$p.value[,2]
limmaRvC.pval<-limma.de.test$p.value[,3]

# weighted limma

HCR.ribo.batch.corrected<-batch.removed.limma.weighted[,1:16]
limma.wt.lm.fit<-lmFit(HCR.ribo.batch.corrected,design = model.matrix(~ species.label))
limma.wtHvC.effect<-limma.wt.lm.fit$coefficient[,2]
limma.wtRvC.effect<-limma.wt.lm.fit$coefficient[,3]

limma.wt.de.test<-eBayes(limma.wt.lm.fit)

limma.wtHvC.pval<-limma.wt.de.test$p.value[,2]
limma.wtRvC.pval<-limma.wt.de.test$p.value[,3]

# combat parametric

HCR.ribo.batch.corrected<-batch.removed.combat.par[,1:16]

combat.par.lm.fit<-lmFit(HCR.ribo.batch.corrected,design = model.matrix(~ species.label))
combat.parHvC.effect<-combat.par.lm.fit$coefficient[,2]
combat.parRvC.effect<-combat.par.lm.fit$coefficient[,3]

combat.par.de.test<-eBayes(combat.par.lm.fit)

combat.parHvC.pval<-combat.par.de.test$p.value[,2]
combat.parRvC.pval<-combat.par.de.test$p.value[,3]

# combat non-parametric

HCR.ribo.batch.corrected<-batch.removed.combat.nonpar[,1:16]

combat.nonpar.lm.fit<-lmFit(HCR.ribo.batch.corrected,design = model.matrix(~ species.label))
combat.nonparHvC.effect<-combat.nonpar.lm.fit$coefficient[,2]
combat.nonparRvC.effect<-combat.nonpar.lm.fit$coefficient[,3]

combat.nonpar.de.test<-eBayes(combat.nonpar.lm.fit)

combat.nonparHvC.pval<-combat.nonpar.de.test$p.value[,2]
combat.nonparRvC.pval<-combat.nonpar.de.test$p.value[,3]



```


Spearman rank correlation coefficient of test statistics
```{r}
HvsC.pval<-c(
cor(mm4HvC.pval,oriHvC.pval, method = "spearman"),
cor(mm4HvC.pval,limmaHvC.pval, method = "spearman"),
cor(mm4HvC.pval,limma.wtHvC.pval, method = "spearman"),
cor(mm4HvC.pval,combat.parHvC.pval, method = "spearman"),
cor(mm4HvC.pval,combat.nonparHvC.pval, method = "spearman")
)
RvsC.pval<-c(
cor(mm4RvC.pval,oriRvC.pval, method = "spearman"),
cor(mm4RvC.pval,limmaRvC.pval, method = "spearman"),
cor(mm4RvC.pval,limma.wtRvC.pval, method = "spearman"),
cor(mm4RvC.pval,combat.parRvC.pval, method = "spearman"),
cor(mm4RvC.pval,combat.nonparRvC.pval, method = "spearman")
)

pval.cor<-as.data.frame(rbind(HvsC.pval,RvsC.pval))
names(pval.cor)<-c("original","limma","limma.wt","combat.par","combat.nonpar")

# correlation of DE test p values between mm4 and the data specified in column name 
pval.cor

HvsC.beta<-c(
cor(mm4HvC.effect,oriHvC.effect, method = "spearman"),
cor(mm4HvC.effect,limmaHvC.effect, method = "spearman"),
cor(mm4HvC.effect,limma.wtHvC.effect, method = "spearman"),
cor(mm4HvC.effect,combat.parHvC.effect, method = "spearman"),
cor(mm4HvC.effect,combat.nonparHvC.effect, method = "spearman")
)
RvsC.beta<-c(
cor(mm4RvC.effect,oriRvC.effect, method = "spearman"),
cor(mm4RvC.effect,limmaRvC.effect, method = "spearman"),
cor(mm4RvC.effect,limma.wtRvC.effect, method = "spearman"),
cor(mm4RvC.effect,combat.parRvC.effect, method = "spearman"),
cor(mm4RvC.effect,combat.nonparRvC.effect, method = "spearman")
)

beta.cor<-as.data.frame(rbind(HvsC.beta,RvsC.beta))
names(beta.cor)<-c("original","limma","limma.wt","combat.par","combat.nonpar")

# correlation of DE test beta values between mm4 and the data specified in column name 
beta.cor
```


scatter plot to visualize effect of batch correction  
```{r}

#human vs chimp
# beta
ggplot(mapping = aes(x=oriHvC.effect,y=mm4HvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("original")+ylab("mm4")+ggtitle("beta HC")

ggplot(mapping = aes(x=limmaHvC.effect,y=mm4HvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("limma")+ylab("mm4")+ggtitle("beta HC")

ggplot(mapping = aes(x=limma.wtHvC.effect,y=mm4HvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("limma.wt")+ylab("mm4")+ggtitle("beta HC")

ggplot(mapping = aes(x=combat.parHvC.effect,y=mm4HvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("combat.par")+ylab("mm4")+ggtitle("beta HC")

ggplot(mapping = aes(x=combat.nonparHvC.effect,y=mm4HvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("combat.nonpar")+ylab("mm4")+ggtitle("beta HC")


# pvalue
ggplot(mapping = aes(x=-log10(oriHvC.pval),y=-log10(mm4HvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("original")+ylab("mm4")+ggtitle("-log10 pvalue HC")

ggplot(mapping = aes(x=-log10(limmaHvC.pval),y=-log10(mm4HvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("limma")+ylab("mm4")+ggtitle("-log10 pvalue HC")

ggplot(mapping = aes(x=-log10(limma.wtHvC.pval),y=-log10(mm4HvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("limma.wt")+ylab("mm4")+ggtitle("-log10 pvalue HC")

ggplot(mapping = aes(x=-log10(combat.parHvC.pval),y=-log10(mm4HvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("combat.par")+ylab("mm4")+ggtitle("-log10 pvalue HC")

ggplot(mapping = aes(x=-log10(combat.nonparHvC.pval),y=-log10(mm4HvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("combat.nonpar")+ylab("mm4")+ggtitle("-log10 pvalue HC")


#Rhesus vs Chimp

# beta
ggplot(mapping = aes(x=oriRvC.effect,y=mm4RvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("original")+ylab("mm4")+ggtitle("beta RC")

ggplot(mapping = aes(x=limmaRvC.effect,y=mm4RvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("limma")+ylab("mm4")+ggtitle("beta RC")

ggplot(mapping = aes(x=limma.wtRvC.effect,y=mm4RvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("limma.wt")+ylab("mm4")+ggtitle("beta RC")

ggplot(mapping = aes(x=combat.parRvC.effect,y=mm4RvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("combat.par")+ylab("mm4")+ggtitle("beta RC")

ggplot(mapping = aes(x=combat.nonparRvC.effect,y=mm4RvC.effect))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("combat.nonpar")+ylab("mm4")+ggtitle("beta RC")


# pvalue
ggplot(mapping = aes(x=-log10(oriRvC.pval),y=-log10(mm4RvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("original")+ylab("mm4")+ggtitle("-log10 pvalue RC")

ggplot(mapping = aes(x=-log10(limmaRvC.pval),y=-log10(mm4RvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("limma")+ylab("mm4")+ggtitle("-log10 pvalue RC")

ggplot(mapping = aes(x=-log10(limma.wtRvC.pval),y=-log10(mm4RvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("limma.wt")+ylab("mm4")+ggtitle("-log10 pvalue RC")

ggplot(mapping = aes(x=-log10(combat.parRvC.pval),y=-log10(mm4RvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("combat.par")+ylab("mm4")+ggtitle("-log10 pvalue RC")

ggplot(mapping = aes(x=-log10(combat.nonparRvC.pval),y=-log10(mm4RvC.pval)))+geom_point(size=1, alpha=0.3)+geom_abline()+xlab("combat.nonpar")+ylab("mm4")+ggtitle("-log10 pvalue RC")

```

function for roc curve
```{r}
roc.curve<-function(pvalue1,pvalue2,cutoff,breaks,color){
  TPR<-c()
  FPR<-c()
  truePositive <- pvalue1 < cutoff
  trueNegative <- pvalue1 > cutoff
  threshold<-seq(0,1,1/breaks)
  
  for (i in 1:breaks+1){
    Detected <- pvalue2 < threshold[i]
    TPR[i] <- length(which(truePositive & Detected))/length(which(truePositive))
    FPR[i] <- length(which(trueNegative & Detected))/length(which(trueNegative))
    
  }
  
  plot(FPR,TPR,xlim=c(0,1),ylim=c(0,1),col=color,pch=16,cex=0.4)
  
}
```

comparing between methods
```{r}

roc.curve(mm4HvC.pval,oriHvC.pval,0.01,1000,"black")
par(new=TRUE)
roc.curve(mm4HvC.pval,limmaHvC.pval,0.01,1000,"green")
par(new=TRUE)
roc.curve(mm4HvC.pval,limma.wtHvC.pval,0.01,1000,"yellow")
par(new=TRUE)
roc.curve(mm4HvC.pval,combat.parHvC.pval,0.01,1000,"blue")
par(new=TRUE)
roc.curve(mm4HvC.pval,combat.nonparHvC.pval,0.01,1000,"red")
legend("bottomright",c("original","limma","limma.wt","combat.par","combat.nonpar"),col=c("black","green","yellow","blue","red"),pch = 16 ,cex = 0.75, pt.cex = 1,box.lwd = 0, ncol = 2)


roc.curve(mm4RvC.pval,oriRvC.pval,0.01,1000,"black")
par(new=TRUE)
roc.curve(mm4RvC.pval,limmaRvC.pval,0.01,1000,"green")
par(new=TRUE)
roc.curve(mm4RvC.pval,limma.wtRvC.pval,0.01,1000,"yellow")
par(new=TRUE)
roc.curve(mm4RvC.pval,combat.parRvC.pval,0.01,1000,"blue")
par(new=TRUE)
roc.curve(mm4RvC.pval,combat.nonparRvC.pval,0.01,1000,"red")
legend("bottomright",c("ori","lm","lmwt","combat.p","combat.np"),col=c("black","green","yellow","blue","red"), pch = 16 ,cex = 0.75, pt.cex = 1,box.lwd = 0, ncol = 2)
```


Venn diagram ori, mm4, corrected (combat, nonparametric)
```{r}

mm4HvC.top500<-mm4HvC.pval<mm4HvC.pval[order(mm4HvC.pval)[501]]
oriHvC.top500<-oriHvC.pval<oriHvC.pval[order(oriHvC.pval)[501]]
combat.nonparHvC.top500<-combat.nonparHvC.pval<combat.nonparHvC.pval[order(combat.nonparHvC.pval)[501]]
HC.pval.top500<-cbind(mm4HvC.top500,oriHvC.top500,combat.nonparHvC.top500)
colnames(HC.pval.top500)<-c("mm4HC","oriHC","corHC")
vennDiagram(HC.pval.top500)

mm4RvC.top500<-mm4RvC.pval<mm4RvC.pval[order(mm4RvC.pval)[501]]
oriRvC.top500<-oriRvC.pval<oriRvC.pval[order(oriRvC.pval)[501]]
combat.nonparRvC.top500<-combat.nonparRvC.pval<combat.nonparRvC.pval[order(combat.nonparRvC.pval)[501]]
RC.pval.top500<-cbind(mm4RvC.top500,oriRvC.top500,combat.nonparRvC.top500)
colnames(RC.pval.top500)<-c("mm4RC","oriRC","corRC")
vennDiagram(RC.pval.top500)


mm4HvC.top1000<-mm4HvC.pval<mm4HvC.pval[order(mm4HvC.pval)[1001]]
oriHvC.top1000<-oriHvC.pval<oriHvC.pval[order(oriHvC.pval)[1001]]
combat.nonparHvC.top1000<-combat.nonparHvC.pval<combat.nonparHvC.pval[order(combat.nonparHvC.pval)[1001]]
HC.pval.top1000<-cbind(mm4HvC.top1000,oriHvC.top1000,combat.nonparHvC.top1000)
colnames(HC.pval.top1000)<-c("mm4","original","corrected")
vennDiagram(HC.pval.top1000)

mm4RvC.top1000<-mm4RvC.pval<mm4RvC.pval[order(mm4RvC.pval)[1001]]
oriRvC.top1000<-oriRvC.pval<oriRvC.pval[order(oriRvC.pval)[1001]]
combat.nonparRvC.top1000<-combat.nonparRvC.pval<combat.nonparRvC.pval[order(combat.nonparRvC.pval)[1001]]
RC.pval.top1000<-cbind(mm4RvC.top1000,oriRvC.top1000,combat.nonparRvC.top1000)
colnames(RC.pval.top1000)<-c("mm4RC","oriRC","corRC")
vennDiagram(RC.pval.top1000)

```

